PRMT1-SFPQ regulates intron retention to control matrix gene expression during craniofacial development

Version 1

Abstract

Abstract

Spliceosomopathies, which are a group of disorders caused by defects in the splicing machinery, frequently affect the craniofacial skeleton and limb, but the molecular mechanism underlying this tissue-specific sensitivity remains unclear. Splicing factors and small nuclear ribonucleoproteins (snRNPs) are core components of splicing machinery, and splicing factors are further controlled by post-translational modifications, among which arginine methylation is one of the most frequent modifications. To determine the splicing mechanisms in cranial neural crest cells (CNCCs), which give rise to the majority of the craniofacial skeleton, we focused on upstream regulators for splicing proteins responsible for arginine methylation, protein arginine methyltransferases (PRMT). These enzymes catalyze arginine methylation of splicing factors to modify splicing factor expression and activity, influencing the splicing product. PRMT1 is the highest expressing enzyme of the PRMT family in CNCCs and its role in craniofacial development is evident based on our earlier investigation, where CNCC-specific Prmt1 deletion caused cleft palate and mandibular hypoplasia. In the present study, we uncover the roles of PRMT1 in CNCCs in the regulation of intron retention, a type of alternative splicing where introns are retained in the mature mRNA sequence. Mandibular primordium of Prmt1-deficient embryos demonstrated an increase in the percentage of intron-retaining mRNA of matrix genes, which triggered NMD, causing a reduction in matrix transcript expression. We further identified SFPQ as a substrate of PRMT1 that depends on PRMT1 for arginine methylation and protein expression in the developing craniofacial structures. Depletion of SFPQ in CNCCs phenocopied PRMT1 deletion in that matrix, Wnt signaling components and neuronal gene transcripts contained higher IR and exhibited lower expression. We further recognized gene length as a common feature among SFPQ-regulated genes in CNCCs. Altogether, these findings demonstrate that the PRMT1-SFPQ pathway modulates matrix Wnt signaling components and neuronal gene expression via intron retention in CNCCs during craniofacial development.

Introduction

Craniofacial abnormalities affecting bone formation in the skull and face are the most common birth defects in infants. Proper formation of these structures involves coordinated processes including proliferation, migration, and differentiation of cranial neural crest cells (CNCCs) (Martik and Bronner 2017; Plein, Fantin, and Ruhrberg 2015). CNCCs are a transient population of progenitor cells that populate the first and second pharyngeal arches and give rise to the maxilla, mandible, palates, and other structures of the developing head (Chai et al. 2000; Jiang et al. 2002; Martik and Bronner 2017). Dysregulation of transcription factors, chromatin remodelers, RNA regulatory proteins, and signaling molecules have been implicated in impaired neural crest development that results in craniofacial defects (Bélanger et al. 2018; Martik and Bronner 2017; Plein, Fantin, and Ruhrberg 2015; Strobl-Mazzulla, Marini, and Buzzi 2012). Spliceosomopathies, which are a group of disorders caused by defects in the splicing machinery, frequently affect the craniofacial skeleton and limb. However, why these tissues are more sensitive to spliceosomal defects remain incompletely understood. Splicing factors and small nuclear ribonucleoproteins (snRNPs) are core components of splicing machinery, which control pre-mRNA splicing and mRNA maturation through the process of alternative splicing (AS). AS includes seven types of events, exon skipping, mutually exclusive exons, alternative 5’ and 3’-splice sites, alternative promoters, alternative polyadenylation, and intron retention (Hooper et al. 2020). Intron retention (IR) is a type of AS abundant in plants and fungi, where introns are retained within the mature mRNA sequence (E.T. Wang et al. 2008). Although initially overlooked in mammals due to low occurrence and challenges in detection, recent technical advances have unveiled its significant roles in various aspects, including cell differentiation and stress response, especially in neuronal, immune, and cancer cells (Marquez et al. 2012; Monteuuis et al. 2019; Braunschweig et al. 2014; Wong and Schmitz 2022).

In exploring the splicing mechanisms in CNCCs, we focused on upstream regulators for splicing factors. The activity of splicing factors is tightly controlled by post-translational modifications, among which methylation is the most abundant, surpassing phosphorylation (Thandapani et al. 2013; Ruta, Pagliarini, and Sette 2021). Splicing regulator methylation mostly occurs on arginine motifs, catalyzed by the protein arginine methyltransferase (PRMT) family of enzymes. PRMTs-mediated methylation of splicing factors determines their subcellular localization, expression, and activity, influencing the splicing product (Blanc and Richard 2017; Liu and Dreyfuss 1995; Snijders et al. 2015). PRMT1 is the most abundant PRMT in most cell types and specifically catalyzes methylation of arginine (R) residues within RG/RGG/GAR repeats, which are motifs enriched in RNA-binding proteins, particularly splicing regulators, resulting in asymmetric dimethylation of arginine (ADMA) (Thandapani et al. 2013; Smith et al. 2004). The importance of PRMT1 in craniofacial development is evident from our earlier investigation in which genetic deletion of Prmt1 in CNCCs leads to cleft palate and mandibular hypoplasia (Gou, Li, Wu, et al. 2018; Gou, Li, Jackson-Weaver, et al. 2018). In the present study, we uncovered a previously unrecognized role of PRMT1 in regulating intron retention (IR) in CNCCs. The mandibular primordium of Prmt1-deficient embryos demonstrated augmented retention of introns in mRNAs that encode bone and cartilage matrix components. We further identified SFPQ as a downstream mediator of PRMT1 in the developing craniofacial structures, that depends on PRMT1 for arginine methylation and protein expression. Depletion of SFPQ in CNCCs partially phenocopied PRMT1 deletion, causing elevated retention of introns in mRNA transcripts of matrix, neuronal, and Wnt signaling pathways genes. We further recognized long gene/intron splicing as a common feature among SFPQ-regulated introns and demonstrated these retained introns contain premature termination codons (PTCs) that trigger nonsense-mediate decay (NMD) to reduce mRNA expression levels. Together, these findings demonstrate that the PRMT1-SFPQ pathway modulates CNCC gene expression via control of splicing. Deficiency of this pathway leads to aberrant intron retention, which induces mRNA decay that downregulates matrix expression in CNCCs during craniofacial development.

Results

CNCCs from the embryonic mandibular process exhibit abundant intron retention, which was further elevated by loss of Prmt1

PRMT1 is the most abundant enzyme of the arginine methyltransferase family in cranial neural crest cells (CNCCs), with an expression level 10∼100 folds higher than other PRMTs (Figure-1A). We previously demonstrated critical roles for PRMT1 in CNCCs during craniofacial skeleton formation using genetic deletion, whereby neural crest-specific deletion of Prmt1 resulted in cleft palate and shorter mandibles (Gou, Li, Wu, et al. 2018; Gou, Li, Jackson-Weaver, et al. 2018). Shorter mandibles, or mandibular hypoplasia is a craniofacial phenotype shared by most spliceosomopathies. To investigate the molecular mechanisms underlying PRMT1’s role in mandible development, we conducted a genome-wide analysis of Prmt1 deletion-induced transcriptional and splicing alterations within the embryonic mandibles. For this purpose, CNCCs, the cell lineage that gives rise to the bone and cartilages in the mandible were isolated from mandibular processes of control (Wnt1-Cre; RosaTd) or Prmt1 CKO mutant (Wnt1-Cre; Prmt1fl/fl; RosaTd) mice at embryonic day 13.5 (E13.5) for mRNA extraction and sequencing (Figure-1B). The ensuing comparative analysis using rMATS (Y. Wang et al. 2024) unveiled changes across five types of alternative splicing events (Figure-1C), echoing our earlier finding where PRMT1 regulates exon usage during cardiac morphogenesis (Jackson-Weaver et al. 2020). In contrast to embryonic epicardial cells, we noted a higher abundance of altered intron retention (IR) in mandibular CNCCs (Figure-1C), with a predominance of IR increase when compared to control CNCCs (Figure-1Ca’). The IR changes were further exemplified by the elevation of intronic expression in Pex12, Nkx3-2, Nkiras1, 1810026B05Rik, and the reduction of intronic expression in Tbx1 (Figure-1D-F). To analyze the intronic expression comprehensively at a gene-specific level, we employed IRI (intron reads index) analysis, which is an algorithm that quantifies the ratio of normalized read counts between intronic and exonic regions (Ni et al. 2016; Sun et al. 2023; Tian et al. 2020). Subsequent analysis for retained introns demonstrated that Prmt1 deficiency-induced change in intron retention altered an array of biology pathways. We noted protein binding, metal ion binding, and cartilage development among genes that present increased intronic expression, while mitochondrion and metabolic pathways were identified among genes with decreased intronic expression (Supplemental Table S1). These findings identify intron retention as a prevalent phenomenon in CNCCs during craniofacial development, and the deletion of Prmt1 further enhanced intron retention in genes involved in diverse biological processes.

CNCC-specific deletion of Prmt1 elevates intron retention in the embryonic mandibular process.

(A) Expression levels of PRMT1-9 mRNAs in primary isolated cranial neural crest cells (CNCC) from Wnt1-Cre; RosaTd mouse embryonic heads at E13.5 and E15.5. TPM, transcript per million. (B) Diagram illustrating the isolation of CNCCs from embryonic mandibles, followed by poly(A)+ mRNA isolation and sequencing (n=4 in control and in Prmt1 CKO). (C) Prmt1 deletion in CNCC caused changes in alternative splicing (AS). Changes in AS events were analyzed by rMATS using RNA sequencing data and significant changes in each type of AS were shown in a stacked bar chart. SE, skipped exon. IR, intron retention. MXE, mutually exclusive exons. A5SS, alternative 5’ splice site. A3SS, alternative 3’ splice site. a) The number of intron retention events that increased (Up, yellow color) or decreased (Down, brown color) in Prmt1 deficient CNCCs. (D & E) Intron retention was prevalent in CNCCs and altered by Prmt1 deletion. Track view of genes demonstrating intron retention and further elevated by Prmt1 deletion, illustrated by Pex12, Nkx3-2, Nkiras1, 1810026B05Rik in D or reduced by Prmt1 deletion, illustrated by Tbx1 in E. (F) Quantification of intron expression in Pex12, Nkx3-2, Nkiras1, 1810026B05Rik and Tbx1 by the percentage of intron-retaining mRNAs in each gene, calculated from IRI analysis based on RNA sequencing data. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; RosaTd. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; RosaTd.

Neural crest-specific deletion of Prmt1 caused a significant reduction of matrix gene expression in the developing mandibles

The main documented roles of retained intron in mature mRNA is to reduce gene expression via non-sense mediated decay (NMD) (Schmitz et al. 2017). When intron-retaining transcripts are incorporated into the translation machinery, introns that contain premature termination codons (PTCs) within their open reading frames trigger NMD, leading to transcript degradation and reduced expression of genes harboring these introns (Wong et al. 2016). We observed that intron-retaining transcripts illustrated in Figure-1, Pex12, and Nkiras1 demonstrated increased intronic expression and decreased exonic expression, suggesting reduced gene expression. To systematically analyze genes that are differentially downregulated, we compared the transcriptional profile between control and Prmt1 CKO mutant of mandibular CNCCs and revealed downregulation of 303 genes and concurrent upregulation of 160 genes by Prmt1 deletion (Figure-2A, 2B). Upon pathway analysis, glycosaminoglycan (GAG) degradation (35 out of 303 genes) and extracellular matrix (ECM) (33 out of 303 genes) emerged as the top pathways among downregulated genes (Figure-2C). Findings from GAG degradation and ECM genes align with our understanding that these downregulated genes encode matrix proteins pivotal to osteogenic and chondrogenic differentiation and the formation of bone and cartilage matrix. This notion was supported by bioinformatic analysis using Ingenuity Pathway Analysis (IPA) based on downregulated genes, which suggested a disruption in bone, cartilage, and connective tissue development within Prmt1 CKO mutant mandibles (Figure-2D). The data collected from transcriptional analysis underscore the importance of PRMT1 in orchestrating matrix gene expression critical to bone, cartilage, and connective tissue formation during mandible morphogenesis.

CNCC-specific Prmt1 deletion reduced matrix gene expression in the developing mandibles.

(A) Volcano plot illustrating upregulation of 160 and downregulation of 303 genes in the mandibular primordium of Prmt1-deficient embryos at E13.5, compared to control embryos. (B) Heatmap showing differential gene expression between control and Prmt1-deficient mandibles. The top downregulated matrix genes were labeled. (C) GO analysis of pathway enrichment in downregulated genes demonstrating glycosaminoglycan (GAG) degradation and extracellular matrix (ECM) organization as the top pathways. (D) Ingenuity Pathway Analysis suggesting connective tissue, bone, and cartilage development as affected biological processes based on downregulated genes. Control: Wnt1-Cre; RosaTd (n=4). Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; RosaTd (n=4).

Deletion of Prmt1 in CNCCs elevated intron retention in mRNAs that encode ECM and GAG degradation enzymes

To determine whether the downregulation of matrix genes is regulated by intron retention, we first examined the intronic expression profile of ECM genes. A prominent elevation of intron retention within ECM genes was revealed by scatter plot where most ECM data points landed above the red line that denotes unchanged intron retention (Figure-3A). ECM genes with the most significant increase in intron expression were labeled and highlighted as red dots, and further exemplified by the track view of Bmp7 and Cthrc1 (Figure-3A, 3B). In these ECM genes, the increased intron retention correlated with the reduced exonic expression, indicated by intronic expression and exonic expression respectively (Figure-3C). To validate the reproducibility of elevated intron retention in ECM transcripts, we expanded our analysis to incorporate four additional embryos from each group, isolated mature mRNA using poly(A)+ magnetic beads, and designed primers for RT-PCR analysis of intronic and exonic region of additional ECM genes. In line with our initial observations, intronic expression increased significantly, coupled with decreased exonic expression in Dcn, Tnn, Lox, Loxl1, Matn2, Col14a1, Adamts12, Adamts2, Adam12, and Scara5 of the mutant CNCCs as compared to the control (Figure-3D). These findings demonstrate that ECM transcripts in Prmt1 CKO mandibles exhibited elevated intron retention, which correlates with a reduction in mRNA expression levels. Next, we analyzed the intronic reads profile of GAG degradation genes, which is the top downregulated gene cluster in the Prmt1 CKO group and clinically associated with craniofacial birth defects. Similar to ECM genes, the majority of GAG degradation genes showed elevation of intron retention indicated by scatter plot (Figure-3E). The increased intron retention of St6galnac3 and Galn11, indicated by higher intronic expression, correlated with reduced mRNA expression, indicated by lower mRNA expression (Figure-3F, 3G). Upon further inspection of the retained introns in ECM and GAG degradation transcripts, retained introns in the significantly downregulated genes all contain PTCs within their open reading frames, suggesting PTC-triggered transcript degradation via NMD-mediated RNA surveillance system (Supplemental Table S2) (Wong et al. 2016). Altogether, these data indicate that Prmt1 deletion elevates intron retention in matrix genes encoding ECM proteins and GAG degradation enzymes and downregulated their mRNA expression within CNCCs.

PRMT1 regulates intron retention in ECM and GAG degradation genes.

(A-C) Intron retention increased in the majority of ECM gene transcripts that were downregulated in Prmt1 deficient embryos, as illustrated by a scatter plot based on intron retention index (IRI) analysis. The red line delineates where intron retention is unchanged. Genes with the top differential IR were represented by red dots and labeled. ECM gene transcripts Bmp7 and Cthrc1 demonstrating higher intron retention in Prmt1- deleted embryos were illustrated by track view in B and quantification of intronic (left) and exonic (right) expression in C. (D) Higher intron retention and lower mRNA level of ECM genes were validated in a panel by RT-PCR. Primers that span the intronic or intron-exon junction region were used to assess intronic expression. Primers that span the exonic region were used to examine exonic expression. (E-G) Intron retention increased in the majority of GAG degradation gene transcripts that were downregulated in Prmt1 deficient embryos compared to control, as indicated by a scatter plot based on IRI analysis. The red line delineates where intron retention is unchanged. Genes were represented by black dots and labeled in red. GAG degradation genes St6galnac3 and Galnt11 demonstrating higher intron retention in Prmt1-deleted embryos were illustrated by (F) track view and (G) quantified for intronic (left) and exonic (right) expression. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; RosaTd. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; RosaTd.

PRMT1 methylates SFPQ in CNCCs

PRMT1 is a methyltransferase that catalyzes methylation of arginine (R) residues within RG/RGG/GAR repeats, which are motifs highly enriched in RNA-binding proteins, particularly splicing regulators (Thandapani et al. 2013). PRMT1 catalyzes asymmetric dimethylation of arginine (ADMA) residues, which regulates the turnover, subcellular localization, and splicing activity of splicing regulators (Smith et al. 2004). We therefore hypothesized that PRMT1 regulates intron retention via methylation of splicing regulators. Earlier studies by Graham et al. and our team have characterized the landscape of arginine methylation governed by PRMT1 and identified a cohort of RNA processing proteins exhibiting PRMT1-dependent methylation (Hartel et al. 2019). Among these, six splicing regulators are highly expressed in embryonic CNCCs: SFPQ, SRSF1, EWSR1, TAF15, TRA2B, and G3BP1 (Figure-4A). To discern whether any of the splicing regulators are accountable for elevated IR in Prmt1 deficient CNCCs, we conducted an in-silico analysis of the RNA-seq data of mandibular CNCCs from control and Prmt1 CKO embryos utilizing rMATS (multivariant analysis of transcript splicing, MATS), followed by rMAPS2 (RNA map analysis) (Hwang et al. 2020; Y. Wang et al. 2024). This integrated approach detects differential splicing events, and then generates visual maps illustrating the spatial distribution of RNA binding protein (RBP) motifs, and finally ranks RBPs based on motif density score and p-value (Park et al. 2016). Through this in-silico analysis, we identified a subset of RBPs with significantly differential binding in splicing events altered by PRMT1 (p<0.00001) (Supplemental Table S3). To pinpoint the most pertinent RBPs, we cross-referenced this significant RBP list with the previously identified six PRMT1-dependent and CNCC-enriched splicing regulators (Figure-4A). SFPQ emerged as the prime candidate, boasting the highest motif scores and the lowest P-values (Supplemental Figure-S1A-E, Supplemental Table S4).

PRMT1 methylates SFPQ and regulates its expression.

(A) Expression levels of splicing factors SFPQ, SRSF1, EWSR1, TAF15, TRA2B, HnRNPA1, WDR70, G3BP1 in primary isolated CNCCs from Wnt1-Cre; RosaTd mouse embryonic heads at E13.5 and E15.5. TPM, transcript per million. (B-F) SFPQ methylation diminished in the mandibular and maxillary processes of Prmt1 deficient embryos at E13.5. B shows that sagittal sections of mouse tissues were used to analyze the mandibular (area outlined in red, bottom) and maxillary (area outlined in red, top) process. Methyl-SFPQ was detected by proximity ligation assay (PLA) in the embryonic mandible (C) and maxilla (E) of control and Prmt1 deficient embryos. Green puncta indicated methyl-SFPQ, and nuclei were counterstained with DAPI (blue). Representative images are shown for Control (Ca-Cd, Ea-Ed) and Prmt1 CKO (Ce-Ch, Ee-Eh). Higher magnification views in (Cd, Ed) and (Ch, Eh) illustrate methyl-SFPQ (green puncta) in CNCC nuclei, as indicated by white arrows. D&F show quantification of PLA puncta normalized to cell number in four biological replicates, presented as mean ± SEM. (G) SFPQ methylation signal remained robust in the epithelial region of craniofacial structures, while CNCC-specific deletion of Prmt1 caused a drastic reduction in the craniofacial mesenchyme. Higher magnification of the mandibular tip showing mesenchyme and epithelium outlined with dotted lines in Control (Ga-Gc) and Prmt1 CKO (Gd-Gf). (H-K) SFPQ protein expression was significantly reduced in the mandibular and maxillary processes of Prmt1 deficient embryos. The level of SFPQ protein was detected by immunostaining in the embryonic mandible (H) and maxilla (J) of control and Prmt1 deficient embryos, and quantified in I and K. (L-M) SFPQ protein expression declined dramatically in the Prmt1 deficient embryonic head, as detected by Western blotting (L). Quantification was conducted with ImageJ and shown in M. Mx, maxilla. Md, mandible. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; RosaTd. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; RosaTd.

Subsequently, we set out to determine whether SFPQ is methylated by PRMT1 in the craniofacial structures. Employing the proximity ligation assay (PLA), a technique designed for the in-situ detection of protein modifications, we probed the E13.5 embryo sections with anti-SFPQ and anti-methyl-arginine ADMA antibodies utilizing the PLA kit, which revealed the presence of methylated SFPQ as distinct green punctate staining. Within the craniofacial complex, a substantial presence of SFPQ methylation was observed in the mandibular processes at E13.5 (Figure-4B, 4Ca-d). In Prmt1 CKO embryos, the signal of methyl-SFPQ showed marked diminishment within the CNCC-derived mesenchyme region (Figure-4Ce-h, 4D). The same analysis was conducted in the maxillary regions, which demonstrated robust SFPQ methylation in the control group and a dramatic reduction of methylation in Prmt1 CKO embryos (Figure-4B, 4Ea-h, 4F). This CNCC-specific deletion of PRMT1 only reduced SFPQ methylation in the CNCCs, as non-CNCC lineages within the epithelium continued to display methyl-SFPQ signals (Figure-4Ga-f, Supplemental Figure-S2). These findings establish SFPQ’s methylation within the craniofacial complex as a PRMT1-dependent phenomenon.

PRMT1-induced methylation controls the activity of splicing regulators by modifying their stability, subcellular distribution, or RNA binding ability (Rho et al. 2007). We then proceeded to examine whether SFPQ protein expression was regulated by PRMT1 in the craniofacial complex. In control embryos, SFPQ exhibited prominent expression in CNCCs and mainly localized in the nucleus (Figure-4Ha-c, 4Ja-c). In Prmt1-deficient CNCCs, the protein’s expression level witnessed a significant reduction within the mandibular and maxillary processes (Figure-4H-4K). A corroborating Western blot (WB) analysis using tissues from the craniofacial structures, including facial and anterior skull regions of embryonic heads, confirmed this decline in SFPQ expression (Figure-4L, 4M). Taken together, these data demonstrate that PRMT1 methylates SFPQ and promotes SFPQ protein expression in the craniofacial complex.

PRMT1-SFPQ pathway regulates the splicing of matrix genes in CNCCs

To investigate the role of SFPQ in intron retention and gene expression within CNCCs, we isolated primary CNCCs from control (Wnt1-Cre; Td) embryos at E13.5 and enriched CNCCs by Td+ signal using cell sorting. Within a day, Td+ CNCCs were transfected with control or two independent SFPQ siRNAs to deplete SFPQ, and then poly(A)+ mRNA was purified followed by sequencing. To determine whether these primary isolated CNCCs retained CNCC characteristics after siRNA-mediated knockdown, the expression of CNCC markers Twist1, Sox10, Msx1, Snai2, and Tfap2a was examined (Achilleos and Trainor 2012; Ishii et al. 2012), which demonstrates that the CNCC marker expression in siRNA-transfected CNCCs after isolation from E13.5 embryos is comparable to that in fresh isolated CNCCs from E13.5 or E15.5 embryos (Supplemental Figure-S3), validating CNCC identity and supporting this method as a feasible approach for mechanistic study of CNCCs using developmental stage-specific embryos. Subsequently, using sequencing data from these siRNA-transfected CNCCs, we conducted bioinformatic analyses of differentially retained introns and differentially expressed genes between control and Sfpq-depleted CNCCs (Figure-5A, 5B). Genes with differential intron retention were assessed by IRI analysis, which revealed elevated intron retention in 397 (siSFPQ #1 vs. siCont) or 388 (siSFPQ #2 vs. siCont) genes and reduced intron retention in 129 (siSFPQ #1 vs. siCont) or 151 (siSFPQ #2 vs. siCont) genes caused by Sfpq depletion (Figure-5A). Genes showing increased IR in Sfpq depleted CNCCs encompassed “proteoglycan metabolic process” and “proteoglycan biosynthesis process”, which included GAG degradation genes, and “connective tissue development”, which includes ECM genes (Figure-5Ca, 5Cb). Sfpq depletion also led to downregulation of 269 (siSFPQ #1 vs. siCont) or 232 (siSFPQ #2 vs. siCont) genes, among which 81 (siSFPQ #1 vs. siCont) and 60 (siSFPQ #2 vs. siCont) genes overlapped with elevated IR, respectively (Figure-5B). To determine intron retention co-regulated by SFPQ and PRMT1, we cross-analyzed genes with increased intronic expression in Sfpq-depleted CNCCs and Prmt1-deficient CNCCs and uncovered a partial overlap of a total of 64 genes (Figure-5D), represented by matrix genes Col4a2, Adam12, Ntn1, App, St6galnac3, Galnt11 Galnt10, and Asph (Figure-5E, 5F). The elevation of intronic retention in these genes also correlated with a reduction of gene expression (Figure-5G, 5H). GO analysis of differentially downregulated genes by both SFPQ-specific siRNAs (cluster 4 within the heatmap) further confirmed extracellular matrix, extracellular structure, and adhesion as the top categories (Figure-5I, 5J). These data indicate that depletion of SFPQ is sufficient to enhance IR and reduce gene expression for a cluster of matrix genes, suggesting loss of SFPQ as a regulatory mechanism for these genes in Prmt1-deficient CNCCs and SFPQ as a downstream effector for PRMT1 in CNCCs.

PRMT1-SFPQ pathway regulates matrix genes in CNCCs.

(A) SFPQ depletion in CNCCs altered intron retention. IR events in genes showing increased (Up, yellow color) or decreased (Down, brown color) intron retention were illustrated by stacked bar graphs. CNCCs were transfected with two independent siRNAs targeting SFPQ, or control siRNA, followed by poly(A)+ mRNA extraction and RNA sequencing. (B) SFPQ depletion in CNCCs altered gene expression. Pie chart demonstrating differentially regulated genes in CNCCs with SFPQ depletion by two independent siRNAs. Shaded areas among the downregulated genes (orange) indicate overlap with IRI up. (C) GO analysis of genes with increased IR events in CNCCs with SFPQ depletion by siRNA#1 (a) or #2 (b). (D) Genes with increased IR events in SFPQ-depleted CNCCs demonstrated 8.28% (64 out of 773) overlap with Prmt1 deficient CNCCs. (E-H) Matrix genes regulated by both Prmt1 deletion and SFPQ depletion. Bar graph showing elevated IR (E, F) and reduced expression (G, H) in both SFPQ-depleted and Prmt1 deletion CNCCs. TPM, transcripts per million. (I-J) Heatmap and GO analysis of SFPQ-regulated genes in CNCC. * p<0.05 siSFPQ vs siControl. siControl, CNCCs transfected with control siRNA. siSFPQ#1 and siSFPQ#2, CNCCs transfected with two independent SFPQ siRNAs. * p<0.05 Prmt1 CKO vs Control. Control, Wnt1-Cre; RosaTd. Prmt1 CKO, Wnt1-Cre; Prmt1fl/fl; RosaTd.

SFPQ depletion in CNCC enhanced intron retention in Wnt signaling and neuronal transcripts and reduced their mRNA expression

More intriguingly, Wnt signaling emerged as the top pathway regulated by SFPQ in CNCCs (Figure-5C). Sfpq depletion increased the percentage of intron-retaining mRNAs of genes in this cluster and reduced total mRNA expression (Figure-6A). These Wnt signaling components-regulated by SFPQ encode both positive and negative regulators of Wnt signaling and encompass both canonical and non-canonical Wnt signal regulators. The SFPQ-regulated Wnt signaling components Wwox also exhibited increased IR and decreased expression in Prmt1 CKO CNCCs. Additionally, neuronal genes occupy a big fraction among SFPQ-regulated introns and genes in CNCCs, as “axon guidance”, “axonogenesis”, “neuron projection development”, and “leaning or memory” emerged as the top biological processes among differentially upregulated introns (Figure-5Ca, 5Cb), and “regulation of membrane potential”, “learning or memory” “cognition” and “locomotory behavior” denotes differentially downregulated genes in Sfpq-depleted CNCCs (Figure-5I, 5J). In CNCCs, Sfpq depletion increased the percentage of intron-retaining mRNAs of these neuronal genes and reduced total mRNA expression (Figure-6B). Upon further inspection, “regulation of membrane potential” was also among identified among genes with decreased expression (38 out of 303 genes) in Prmt1 CKO CNCCs. Altogether, these findings demonstrate that, besides matrix genes, SFPQ regulates Wnt signaling components and neuronal gene expression via intron retention.

SFPQ regulates intron retention of Wnt signaling and neuronal genes in CNCCs.

SFPQ depletion in CNCCs elevated intron retention and decreased mRNA expression of a group of Wnt signaling components (A) and neuronal genes (B). The levels of mRNA expression (left) and intron retention (right) were illustrated by a two-sided bar graph. TPM, transcripts per million. * p<0.01 siSFPQ vs siControl. siControl, CNCCs transfected with control siRNA. siSFPQ#1 and siSFPQ#2, CNCCs transfected with two independent SFPQ siRNAs.

Disturbance of SFPQ activity impaired splicing of long genes and long introns

SFPQ predominantly binds to intronic regions and regulates splicing (Hosokawa et al. 2019). We therefore used published CLIP-seq dataset generated from embryonic mouse brain to determine whether these aberrantly retained introns are SFPQ targets (Hosokawa et al. 2019). The retained Ptk7 intron 1 encompassed a much higher density of SFPQ binding peaks when compared to the spliced intron 6 within the same gene (Figure-7A). We further conducted analysis for all the introns at the genomic level comparing SFPQ binding peaks among introns with elevated retention to introns with reduced retention or no changes and demonstrated that introns with elevated retention in Sfpq-depleted CNCCs showed significantly higher enrichment of SFPQ binding peaks (Figure-7B). We further analyzed SFPQ CLIP-seq peaks at the gene level, comparing genes with elevated IR to genes with no IR changes. Our analysis showed that genes containing elevated IR in Sfpq-depleted CNCCs exhibited a trend towards higher presence of SFPQ binding peaks based on this embryonic brain dataset (Figure-7C). These data suggest that introns with aberrant IR elevation in Sfpq-depleted CNCCs are direct splicing targets bound by SFPQ.

SFPQ depletion reduces long gene expression via intron retention triggered NMD.

(A) SFPQ binding peaks were mapped to retained intron 1, but not spliced intron 6 of Ptk7. Peak distribution from published Sfpq CLIP-seq data using E13.5 brain (top) and track view of RNA-seq data using siCont or siSFPQ-transfected CNCCs (bottom) for Ptk7, with the retained Ptk7 intron 1 and spliced intron 6 highlighted. (B) SFPQ binding peaks were significantly enriched in retained introns within CNCCs. Violin plot displaying the global distribution of enriched SFPQ binding peaks in introns with elevated retention compared to introns with reduced retention or no change. Peak number was normalized to intron length. P value was indicated at the top. (C) SFPQ binding peaks were preferentially enriched in genes with higher intron retention when compared to genes with no IR change. p = 0.07 using Fisher’s exact test. (E) SFPQ-regulated genes were significantly longer. Violin plot displaying the distribution of length for genes showing increased intron retention (IR Up), decreased intron retention (IR Down), or unchanged intron expression (No Change). (F) SFPQ depletion promoted intron retention and reduced mRNA expression of Col4a2, St6galnac3, and Ptk7 in ST2 cells. Bar chart showing RT-PCR analysis of intronic and exonic expression. * P<0.05 siSFPQ vs. siControl. (G-H) NMD inhibitor NMDI14 induced the accumulation of retained introns and mRNAs of Col4a2, St6galnac3, and Ptk7 in ST2 cells. Bar chart showing RT-PCR analysis of intronic (G) and exonic expression (H) in DMSO or NMDI-treated cells. * P<0.05 NMDI vs. DMSO treatment of the same group.

To understand the molecular basis of how SFPQ targets these matrix, Wnt and neuronal genes and introns, we evaluated characteristics shared among SFPQ-regulated genes, and recognized that they are either long genes (>100kd in length) or retain long introns (>10kd in length). Of note, Wnt pathway gene Wwox and matrix genes St6galnac3 are 913kd and 526kd in length, and their retained introns are extremely long, 639kd and 215kd, respectively, in contrast to a median intron size of 0.6 ∼ 2.4kb in the mouse genome (Hong, Scofield, and Lynch 2006). In the human and mouse genome, long genes mostly occur because of long introns (Breschi, Gingeras, and Guigó 2017; Lopes et al. 2021). We then plotted the length of SFPQ-regulated genes against genes across the genome and uncovered that the genes with increased IR in Sfpq depleted CNCCs have a median length of ∼100kb, which is much longer than genes in the control groups, where the medial length is around 10kb (Figure-7D). These findings suggest that SFPQ-regulated splicing facilitates regulation of long gene expression during craniofacial development.

SFPQ depletion reduces matrix and Wnt signaling gene expression via IR-induced NMD

In Sfpq-depleted CNCCs, retained introns contain PTCs within their open reading frames, suggesting PTC-triggered NMD as the mechanism of degradation (Supplemental Table S5). To test intron-regulated mRNA decay, we utilized the ST2 cell line, which is a mesenchymal cell line frequently used in the mechanistic study of ossification. SFPQ-regulated intron splicing and mRNA expression for Col4a2, St6galnac3, and Ptk7 is conserved in ST2 cells where Sfpq depletion caused an increase in the percentage of intron-retaining mRNAs, and a concurrent decrease in their transcript levels (Figure-7E). To determine whether Col4a2, St6galnac3, and Ptk7 mRNA were degraded through intron-induced NMD, we inhibited NMD with a chemical inhibitor, NMDI14, which targets key components of the NMD machinery, Upf1, and Smg5/7 (Martin et al. 2014). NMDI14 induced accumulation of intron 4 of Col4a2, intron 1 of St6galnac3, and intron 1 of Ptk7 (Figure-7F), indicating that mRNA decay of intron-retaining transcripts occurs through the NMD machinery. To assess the impact of NMD on the total level of these mRNAs, we used PCR primers that detect the fully spliced transcripts and observed that inhibition of NMD restored and increased the levels of these mRNAs in Sfpq-depleted ST2 cells (Figure-7G). These data indicate that depletion of Sfpq causes aberrant intron retention that accelerates mRNA decay through NMD.

Discussion

Here we revealed prevalent intron retention within CNCCs during embryonic development and propose intron-regulated expression of long genes as a causal mechanism for craniofacial deformity. This study demonstrated PRMT1-SFPQ pathway-regulated intron retention and subsequent control of matrix and Wnt pathway component gene expression during craniofacial development. Genetic deletion of Prmt1 or depletion of Sfpq in CNCCs caused aberrant intron retention in genes encoding matrix proteins and Wnt pathway components, which triggers NMD to degrade these transcripts. Matrix and Wnt pathway genes regulated by this mechanism share a common feature as long genes with a median length of 100kb, suggesting higher susceptibility of these genes to perturbation of the PRMT1-SFPQ pathway or splicing defects during craniofacial development. IR has been linked to diseases like cancer and autoimmune diseases (Dvinge and Bradley 2015). This study on PRMT1-regulated IR in controlling the expression of matrix genes in CNCCs, to our knowledge, is the first characterization of IR in craniofacial development. Many of the PRMT1-SFPQ pathway regulated matrix and Wnt signaling genes are linked to congenital conditions. For example, in our study, the long gene Wwox exhibited increased retention of long introns (introns 3 and 4) and decreased expression upon Prmt1/Sfpq deficiency. In human patients, pathogenic variants of WWOX with large deletion within long introns have been associated with epileptic encephalopathy syndrome manifesting shared facial phenotype (Dvinge and Bradley 2015). This deletion within these long introns may affect SFPQ binding or splicing that impairs PRMT1-SFPQ pathway-regulated splicing of human WWOX genes. Another SFPQ target, Ptk7 is associated with neural tube defects and scoliosis (Berger, Wodarz, and Borchers 2017). SFPQ itself is linked to neurodegenerative diseases including amyotrophic lateral sclerosis, frontotemporal dementia, and Alzheimer’s disease, and was recently identified by a transcriptional study as a genetic regulator in CNCCs (Feng et al. 2021).

A main feature shared by SFPQ-regulated genes in CNCCs is gene length, with a median length of 100kb in contrast to a median length of ∼10kb across the genome. In each SFPQ-regulate gene, introns exhibiting the highest retention also tend to be longer introns, for example, intron 4 of Col4a2 which displayed the most dramatic increase in intronic reads within Col4a2 boasts a length of 45kb. Increased IR in neuronal and immune cells has been proposed to facilitate a rapid response to external stimuli, allowing a time frame shorter than that required for de novo transcription (Ni et al. 2016; Mauger, Lemoine, and Scheiffele 2016). Thus, during craniofacial morphogenesis, IR represents a developmental process that enables rapid production of proteins when they are required. Long genes over 100kb in length require two to several hours for transcription, which presents a challenge for the fast tempo during embryogenesis. This intron-regulated mechanism allows cells to control the rate of production post-transcriptionally, and to achieve rapid protein translation without waiting for the time period required for transcription. Long genes regulated by this mechanism are highly enriched in matrix pathways, and responsible for the production of building blocks for bone, cartilage, and connective tissues. Since spliceosomopathies preferentially affect the craniofacial skeleton and limb, intron retention-regulated matrix expression provides mechanistic insights for the higher susceptibility of these tissue types to spliceosome dysfunction.

PRMT1 is a critical regulator of RNA splicing. In mammalian cells, it methylates over forty splicing factors to modulate protein stability and subcellular localization, thereby influencing their accessibility to mRNA (Thandapani et al. 2013; Wu et al. 2022). PRMT1-catalyzed arginine methylation also affects their RNA binding affinity and specificity, altering their activity and the splicing product (Fedoriw et al. 2019; Fong et al. 2019). Therefore, it is not surprising that the disruption of this methyltransferase leads to splicing defects. However, the elevation of IR in this study presented a distinct role for PRMT1 in CNCCs. IR has been linked to diseases like cancer and autoimmune diseases (Dvinge and Bradley 2015). This study on PRMT1-regulated IR in controlling the expression of matrix genes in CNCCs, to our knowledge, is the first characterization of IR in craniofacial development, adding developmental insights to our knowledge of the roles of IR in pathophysiology. Manipulation of SFPQ expression (knockdown with 50% efficiency) demonstrated that SFPQ shares around 10% overlap with PRMT1 in IR-regulated neural crest gene expression. Given that PRMT1 methylates over 40 splicing factors in mammals and many of these splicing factors are abundantly expressed in embryonic CNCCs, we expect that PRMT1-governed splicing will be mediated by multiple splicing regulators besides SFPQ. The functional network of splicing regulation during craniofacial morphogenesis awaits further investigation.

We recognized SFPQ as a key PRMT1 substrate in CNCCs based on an unbiased in silico analysis and further validation in the mandibular and maxillary primordium. This aligns with previous work demonstrating that PRMT1 is responsible for the association of SFPQ with mRNA in mRNP complexes in mammalian cells (Snijders et al. 2015). Our findings on altered neuronal genes in Sfpq-depleted CNCCs also echoed previous reports in which SFPQ regulates many neuronal genes involved in axon extension, branching, viability, and synaptogenesis via direct binding to their intronic regions (Ruskin, Zamore, and Green 1988; Cosker et al. 2016) and loss of SFPQ function compromises splicing patterns, especially the accurate splicing of long introns in embryonic brain and AML samples (Luisier et al. 2018; Taylor et al. 2022; Thomas-Jinu et al. 2017). Findings in this study further revealed previously unappreciated roles of SFPQ in the regulation of matrix genes through IR, which bear functional significance in CNCCs and craniofacial skeleton formation.

Altogether, these findings demonstrate that the PRMT1-SFPQ pathway regulates intron retention in CNCCs and suggests intron retention in long introns or long genes as a mechanism that controls matrix and Wnt signaling gene expression during craniofacial morphogenesis.

Material and methods

Animals

Prmt1fl/fl mice were generously provided by Dr. Stephane Richard (McGill University). Wnt1-Cre; Prmt1fl/fl; R26R tdTomato and Wnt1-Cre; R26RtdTomato mice were obtained by mating Wnt1-Cre, R26RtdTomato mice (Jax #009107 and #007914) with Prmt1fl/fl mice. Animals were genotyped using established protocols (Chai et al., 2000; Yu et al., 2009) and all care and experiments followed USC’s IACUC protocols.

Proximity Ligation Assay (PLA)

PLA was performed using a Duolink In Situ kit (Millipore Sigma Cat# DUO92101-1KT), SFPQ antibody (Everest Biotech Cat#EB09523), and pan-methyl arginine antibody, Asymmetric Di-Methyl Arginine (ADMA) (Cell Signaling Technology Cat# 13522) to detect methyl-SFPQ. Tissue sections were blocked with the Duolink blocking solution in a humidity chamber for 1 hour at 37°C before incubating with the primary antibodies (diluted 1:50 in Duolink PLA probe diluent) overnight at 4°C. The next day, tissue sections were incubated with the Goat-PLUS and Rabbit-MINUS PLA probes for 1 hour at 37°C, followed by ligase and amplification solution. After the final washes, tissue sections were mounted with Duolink In Situ mounting media with DAPI to counterstain nuclei. Images were visualized under the Confocal microscope. The number of PLA signals that appear as punctuate green signals in the nuclei of the cells and the quantification was performed with CellProfille software.

Tissue Processing and Immunofluorescence Staining

E13.5 control and Wnt1-Cre; Prmt1fl/fl mice embryos were fixed in 4% PFA overnight, dehydrated, embedded in O.C.T., sectioned at 8 µm thickness in sagittal orientation, and mounted on glass slides. The samples were submitted to antigen retrieval using Antigen Unmasking Solutions (Vector Laboratories Cat# H-3300-250) and washed and incubated in 0.5% Triton X-100 in PBS for 20 minutes for permeabilization. The samples were briefly washed with PBS, blocked with 10% goat serum for 1 hour, and then incubated with primary antibody SFPQ (Cell Signaling Technology Cat#23020, 1:100) at 4°C overnight in a moisture chamber. After this, the samples were blocked with a secondary antibody (1:400, goat anti-rabbit IgG (H+L) Alexa Fluor 488 Cat# A-11008) for 1 hour followed by DAPI diluted 1:1000 in PBS for 10 minutes. Finally, the samples were mounted with mounting media (Electron Microscopy Science Cat# 1798510) and covered with cover glass for imaging using Keyence BZ-X800 and Leica DMI 3000B microscopes and quantification with CellProfille software.

Western Blotting

Heads of E13.5 control and Wnt1-Cre; Prmt1fl/fl mice were dissected, followed by removal of the brain. The tissue was homogenized with a pestle followed by lysis with RIPA Lysis and Extraction Buffer. The total protein concentration was determined by comparison with BSA standards. Twenty micrograms of total protein from each sample were loaded into each well of a 10% polyacrylamide gel. Western blot analysis was carried out as previously described (Iwata et al., 2012). Antibodies against PRMT1 (Cell Signaling, Cat# 2449; 1:1000), SFPQ (Cell Signaling Technology Cat#23020, 1:1000), and GAPDH (Cell Signaling, Cat# 97166; 1:5000) were used for Western blot. The samples were analyzed three times in independent blots and the bands were quantified by optical densitometry. The analysis was performed using ImageJ digital imaging processing software (ImageJ 1.48v, National Institutes of Health, Bethesda, MD, USA). The expression of each analyzed protein was normalized with GAPDH.

NCC collection and FACS analysis

The heads of E13.5 Wnt1-Cre; R26RtdTomato and Wnt1-Cre; Prmt1fl/fl;R26RtdTomato mice were dissected and placed in a sterile tube containing Hanks’ Balanced Salt Solution (Thermo Fisher Scientific Cat# 14175079). Following centrifugation, the supernatant was discarded, and the tissue was subjected to TrypLE (Thermo Fisher Scientific Cat# 12605010) incubation for 5-15 minutes at 37°C on a rotator. The resulting dissociated tissue was neutralized by the addition of fetal bovine serum (FBS) and then filtered through a 40μm strainer to obtain a single-cell suspension. After centrifugation, cells were resuspended in an appropriate volume of serum-free medium for FACS analysis. Sorting was performed based on tdTomato fluorescence (Excitation/emission: 554/582 nm), and td-positive cells were collected into separate tubes containing DMEM (Genesee Scientific Cat#25-501) supplemented with 20% FBS.

Primary CNCC culture and siRNA transfection

CNCCs were cultured in DMEM supplemented with 20% FBS in an incubator until they attach to the bottom. Then, reverse transfection with control siRNA (Qiagen Cat# 1027310), SFPQ siRNA #1 (Qiagen Cat#SI05783848), and SFPQ siRNA #2 (Qiagen Cat#SI05783876) at 40 nM using Lipofectamine RNAiMax transfection reagent (Invitrogen) for siRNA delivery were performed. 48h later, total RNA was isolated using TRIzol reagent (Invitrogen) following the manufacturer’s protocols followed by mRNA isolation using NEBNext High-Input Poly(A) mRNA Magnetic Isolation Module (NEBNext® Cat#E3370S).

mRNA isolation and sequencing

Poly(A) mRNA isolation was extracted from total RNA by using the NEBNext High-Input Poly(A) mRNA Magnetic Isolation Module (NEBNext® Cat#E3370S). Five sets of primary isolated CNCCs from control (Wnt1-Cre;ROSATd) or Prmt1 CKO (Wnt1-Cre; Prmt1fl/fl; ROSATd), and three sets of isolated CNCCs transfected with siControl or siSFPQs were sequenced at 40 million reads sequencing depth and 150 bp paired-end sequencing.

Bioinformatic Analysis of Differentially Regulated Genes (DEG) and differential Intron Retention (IR)

The sequencing quality of RNA-seq libraries was assessed by FastQC v0.11.8 (https://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Low-quality bases and adapters were trimmed using Trim Galore (v0.6.7,https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/). The reads were mapped to mouse genome mm10 using hisat2 (v2.1.0) (Kim et al. 2019). The mapped sam files from hisat2 were converted to bam files which were then further turned to sorted bam files by samtools (v1.7) (Li et al. 2009). Mapped reads were then processed by htseq-count (v1.99.2) to calculate the read count of all genes (Anders, Pyl, and Huber 2015). The sorted bam files were then given to IRTools (https://github.com/WeiqunPengLab/IRTools/) to calculate the IR of each gene and the read count of genes’ constitutive intronic regions (CIRs) and constitutive exonic regions (CERs). The expression level of a gene was expressed as TPM (Transcripts Per Kilobase Million) value. EdgeR was used to identify differentially expressed genes (DEGs) by requiring ≥ 1.5-fold expression changes and adjusted p-value < 0.05 (Robinson, McCarthy, and Smyth 2010). The intron retention index (IRI) of a gene is defined as the ratio of the overall read density of its CIRs to the overall read density of its CERs. Genes with 0< IRI <1 and CER expression greater than 1 were selected to identify differential IR genes. Student’s t-test was used to identify differential IR genes by requiring ≥ 1.5-fold log2(IR) changes and p-value < 0.05. Gene Ontology analysis was done by DAVID (https://david.ncifcrf.gov/tools.jsp) and Metascape (https://metascape.org/gp/index.html#/main/step1) (Sherman et al. 2022; Zhou et al. 2019). The RNA-seq data are deposited to the Gene Expression Omnibus (GEO) database with accession number GSE266474.

rMATS and rMAPS Analysis

rMATS (https://github.com/Xinglab/rmats-turbo) and rMAPS2 (http://rmaps.cecsresearch.org/) were used to find potential RNA binding Proteins (RBPs) that contribute to differential splicing events (Hwang et al. 2020; Y. Wang et al. 2024). rMATS was first used to identify differential alternative splicing events between wild type and PRMT1 CKO samples by requiring FDR < 0.05. The output of rMATs were then fed to rMAPS2 to identify significant RBPs contributing to each type of Splicing events.

Quantitative PCR (qPCR)

For both experiments, Control, and Wnt1-Cre; Prmt1fl/fl mandibles, and siRNA-transfected CNCCs, mRNAs were quantified by real-time PCR with IQ Sybr Green Supermix (Bio-Rad) and normalized against GAPDH mRNA levels. Relative changes in expression were calculated using the ΔΔCt method. Primer sequences are listed in Supplemental Table S6.

Statistical analysis

Two-tailed Student’s t-tests or Fisher’s exact tests were applied for statistical analysis. For all graphs, error bars represent standard deviations. A p-value of <0.05 was considered statistically significant.

eLife Assessment

This study presents an interesting mechanism for the regulation of RNA levels, establishing an important regulatory connection between protein arginine methyltransferase 1 and the splicing factor SFPQ. While these findings have theoretical implications beyond a single field, the evidence is incomplete, with only partial support for the main claims.

Reviewer #1 (Public review):

The central finding of the current manuscript is that embryonic ablation of PRMT1 results in a craniofacial phenotype that is primarily linked to downstream intron splicing defects. This manuscript is one of several to underscore the relative importance of intron splicing to gene expression regulation during development, and moreover, to recapitulate splicing-related craniofacial defects. Specifically, authors introduce a regulatory axis consisting of PRMT1-SFPQ that directs mechanisms of long intron retention. This finding represents a significant contribution to our understanding of splicing regulation, in the sense that it highlights the regulatory impact that post-translational modification of splicing-related proteins can have on intron processing. Further, it emphasizes the importance of extending the study of splicing regulation beyond core components of the spliceosome, to include their upstream regulators as well.

The significance of neural crest cells in the development of craniofacial structures has long been considered a major contributor to developmental phenotypes. This specific symptomology is heavily associated with spliceosomopathies, wherein disruption of spliceosome components is the primary mechanism of disease pathogenesis. Thus, the PRMT1 associated phenotype is noteworthy. The role of PRMT1 in methylating downstream splicing factors introduces a new avenue of research focused on the mechanisms of spliceosome component activation and their effects on splicing. The strength of the current study lies in their establishing the molecular mechanism through which PRMT1 could alter craniofacial development through regulation of the transcriptome, but the data presented to support the claim that a PRMT1-SFPQ axis directly regulates intron retention of the relevant gene networks should be robust and with multiple forms of clear validation. For example, elevated intron retention findings are based on the intron retention index, and according to the manuscript, are assessed considering the relative expression of exons and introns from a given transcript. However, delineating between intron retention and other forms of alternative splicing (i.e., cryptic splice site recognition) requires a more comprehensive consideration of the intron splicing defects that could be represented in data. A certain threshold of intron read coverage (i.e., the percent of an intron that is covered by mapped reads) is needed to ascertain if those that are proximal to exons could represent alternative introns ends rather than full intron retention events. In other words, intron retention is a type of alternative splicing that can be difficult to analyze in isolation given the confounding influence of cryptic splicing and cryptic exon inclusion. If other forms of alternative splicing were assessed and not detected, more confident retention calls can be made.

While data presented to support the PRMT1-SFPQ activation axis is quite compelling, that this is directly responsible for the elevated intron retention remains enigmatic. First, in characterizing their PRMT1 knockout model, it is unclear whether the elevated intron retention events directly correspond to downregulated genes. Moreover, intron splicing is a well-documented node for gene regulation during embryogenesis and in other proliferation models, and craniofacial defects are known to be associated with 'spliceosomopathies'. However, reproduction of this phenotype does not suggest that the targets of interest are inherently splicing factors, and a more robust assessment is needed to determine the exact nature of alternative splicing in this system. Because there are several known splicing factors downstream of PRMT1 and presented in the supplemental data, the specific attribution of retention to SFPQ would be additionally served by separating its splicing footprint from that of other factors that are primed to cause alternative splicing.

Clarifying the relationship between SFPQ and splicing regulation is important given that the observed splicing defects are incongruous with published data presented by Takeuchi et al., (2018) regarding SFPQ control of neuronal apoptosis in mice. In this system, SFPQ was more specifically attributed to the regulation of transcription elongation over long introns and its knockout did not result in significant splicing changes. Thus, to establish the specificity for the SFPQ in regulating these retention events, authors would need to show that the same phenotype is not achieved by mis-regulation of other splicing factors. That the authors chose SFPQ based on its binding profile is understandable but potentially confounding given its mechanism of action in transcription of long introns (Takeuchi 2018). Because mechanisms and rates of transcription can influence splicing and exon definition interactions, the role of SFPQ as a transcription elongation factor versus a splicing factor is inadequately disentangled by authors.

Reviewer #2 (Public review):

Summary:

The manuscript by Lima et al examines the role of Prmt1 and SFPQ in craniofacial development. Specifically, the authors test the idea that Prmt1 directly methylates specific proteins that results in intron retention in matrix proteins. The protein SFPQ is methylated by Prmt1 and functions downstream to mediate Prmt1 activity. The genes with retained introns activate the NMD pathway to reduce the RNA levels. This paper describes an interesting mechanism for the regulation of RNA levels during development.

Strengths:

The phenotypes support what the authors claim that Prmt1 is involved in craniofacial development and splicing. The use of state-of-the-art sequencing to determine the specific genes that have intron retention and changes in gene expression is a strength.

Weaknesses:

Some of the data seems to contradict the conclusions. And it is unclear how direct the relationships are between Prmt1 and SFPQ.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 ( Public review):

The strength of the current study lies in their establishing the molecular mechanism through which PRMT1 could alter craniofacial development through regulation of the transcriptome, but the data presented to support the claim that a PRMT1-SFPQ axis directly regulates intron retention of the relevant gene networks should be robust and with multiple forms of clear validation. For example, elevated intron retention findings are based on the intron retention index, and according to the manuscript, are assessed considering the relative expression of exons and introns from a given transcript. However, delineating between intron retention and other forms of alternative splicing (i.e., cryptic splice site recognition) requires a more comprehensive consideration of the intron splicing defects that could be represented in data. A certain threshold of intron read coverage (i.e., the percent of an intron that is covered by mapped reads) is needed to ascertain if those that are proximal to exons could represent alternative introns ends rather than full intron retention events. In other words, intron retention is a type of alternative splicing that can be difficult to analyze in isolation given the confounding influence of cryptic splicing and cryptic exon inclusion. If other forms of alternative splicing were assessed and not detected, more confident retention calls can be made.

This manuscript is a mechanistic exploration that follows previous work we published on the role of Prmt1 in craniofacial development, in which genetic deletion of Prmt1 in CNCCs leads to cleft palate and mandibular hypoplasia (PMID: 29986157).

As the reviewer pointed out, a certain threshold of intron read coverage is needed to assess intron retention events. We employed IRTools to assess the collective changes of intron retention between cell-states associated with certain biological function or pathway. IRTools incorporated considerations for intron read coverage by checking the evenness of read distribution in an intron. Specifically, every constitutive intronic regions (CIR) is divided into 10 equally sized bins and the proportion of reads that map to each bin is calculated. CIRs are then ranked according to their imbalance in bin-wise reads distribution, represented by the proportion of reads in its most populated bin. Those among top 1% are considered to contain potentially false IR events and excluded. We further addressed this question by developing another measure of intron retention, intron retention coefficient (IRC), which assesses IR events using the junction reads (Supplemental Figure-S8). Junction reads that straddle two exons are called exon-exon junction reads (spliced reads), and those that straddle an exon and a neighboring intron are called exon-intron junction reads (retained reads). The IRC of an intron is defined as the fraction of junction reads that are exon-intron junction reads: IRC = exon-intron read-count / (exon-exon read-count + exon-intron read-count), where exon-intron read-count = (5’ exon-intron read-count + 3’ exon-intron read-count) / 2. The IRC of a gene is defined as the exon-intron fraction of all junction reads overlapping or over the constitutive introns of this gene. In the calculation of the IRC, only exon-intron junction reads that cover the junction point and overlap both of each side for at least 8 bps were counted, and only exon-exon junction reads that jump over the relevant junction points and overlap each of the respective exons for at least 8 bps were counted. In this process, evenness of the proportion of exon-intron junction reads that are 5’ or 3’ exon-intron junction reads are taken into account. As shown in the Supplemental Figure S7A and S7B, IRC analysis generated consistent results with those obtained from using IRI (Figure 3A and 3I).

In addition, as the reviewer pointed out, intron retention can be difficult to analyze in isolation. We followed the reviewer’s suggestion that “If other forms of alternative splicing were assessed and not detected, more confident retention calls can be made“ and analyzed other forms of alternative splicing for all ECM and GAG genes with significant IRI increase (genes highlighted in Figure-3A and 3I) using rMATS (Supplemental Figure-S9). Among these genes, only 5 genes (Cthcr1, Mmp23, Adamts10, Ccdc80 and Col25a1) showed statistically significant changes in skipped exon, 1 gene (Bmp7) showed significant changes in mutually exclusive exons, and none showed significant changes in alternative 5’ or 3’ splicing. SE and MXE changes detected were marginal (Supplemental figure S8), while the majority of matrix genes with significant intron retention didn’t exhibit other forms of alternative splicing, further supporting the confidence of intron retention calls.

While data presented to support the PRMT1-SFPQ activation axis is quite compelling, that this is directly responsible for the elevated intron retention remains enigmatic. First, in characterizing their PRMT1 knockout model, it is unclear whether the elevated intron retention events directly correspond to downregulated genes.

In the revised manuscript, we demonstrate IR-triggered NMD as a mechanism for transcript decay and downregulation of matrix genes. When IR-triggered NMD was blocked by chemical inhibitor NMDI14, the intron-retaining transcripts showed significant accumulation (new Figure-4). NMD is the RNA surveillance system to degrade aberrant RNAs. Intron retention-triggered NMD in cancer has both promotive and suppressive roles and NMD inhibitors has been tested for cancer therapy including immunotherapy. During embryonic development, the functional significance of NMD machinery is suggested by human genetic findings and mouse genetic models. NMD is driven by a protein complex composed of SMG and UPF proteins. Smg6, Upf1, Upf2 and Upf3a knockout mouse die at early embryonic stages (E5.5-E9.5), and Smg1 gene trap mutant mice die at E12.5 (PMID: 29272451). SMG9 mutation in human patients causes malformation in the face, hand, heart and brain (PMID: 27018474).

We show that in CNCCs NMD functions both as a physiological mechanism and invoked by molecular insult. Blocking NMD in CNCCs caused significant accumulation of intron-retaining Adamts2, Alpl, Eln, Matn2, Loxl1 and Bgn transcripts, suggesting a basal role for NMD to degrade intron-retaining transcripts (Figure-4Ba-4Bf). We further demonstrated the accumulation of Adamts2 and Fbln5 using semi-quantitative PCR with the detection of a longer product from Adamts2 intron 19 and Fbln5 intron 7 (Figure-4Ca-4Ch). In CNCCs and ST2 cells, NMD is further invoked by Prmt1 and Sfpq deficiency. In Prmt1 deficient CNCCs, NMD blockage led to higher accumulation of intron-retaining Adamts2 and Alpl transcripts, suggesting that Prmt1 deficiency triggers NMD to reduce intron-containing transcripts (Figure-4Aa, 4Ab). In Sfpq-depleted ST2 cells, blocking NMD caused accumulation of intron-retaining transcripts Col4a2, St6galnac3 and Ptk7 (Figure-9B, 9C).

Moreover, intron splicing is a well-documented node for gene regulation during embryogenesis and in other proliferation models, and craniofacial defects are known to be associated with 'spliceosomopathies'. However, reproduction of this phenotype does not suggest that the targets of interest are inherently splicing factors, and a more robust assessment is needed to determine the exact nature of alternative splicing in this system. Because there are several known splicing factors downstream of PRMT1 and presented in the supplemental data, the specific attribution of retention to SFPQ would be additionally served by separating its splicing footprint from that of other factors that are primed to cause alternative splicing.

We have previously shown that a group of splicing factors depends on Prmt1 for arginine methylation, including SFPQ (PMID: 31451547). We tested additional splicing factors that are highly expressed in CNCCs and depends on PRMT1 for arginine methylation: SRSF1, EWSR1, TAF15, TRA2B and G3BP1 (Figure-5, 6 and 10). Among these factors, EWSR1 and TRA2B are both methylated in CNCCs and depend on PRMT1 for methylation (Fig. 5 and Supplemental Figure-S3B, S3C). We weren’t able to assess TAF15 methylation because of lack of efficient antibody for the PLA assay. We also demonstrated that their protein expression or subcellular localization was not altered by Prmt1 deletion in CNCCs, unlike SFPQ (Supplemental Figure-S4). To define their splicing footprint, we performed siRNA-mediated knockdown in ST2 cells, followed by RNA-seq and IRI analysis to define differentially regulated genes and introns, which revealed distinct biological pathways regulated by SFPQ, EWSR1, TRA2B and TAF15, but minimal roles of EWSR1, TRA2B and TAF15 on intron retention when compared to SFPQ (Fig. 10F-10S, Supplemental Figure S7A-S7F, Supplemental Tables S4-S6). ECM genes are significantly downregulated by all four splicing factors (Fig. 10F-10I), but EWSR1, TRA2B and TAF15 function through IR-independent mechanisms, such as exon skipping, as exemplified by Postn (Fig. 10J-10S).

Clarifying the relationship between SFPQ and splicing regulation is important given that the observed splicing defects are incongruous with published data presented by Takeuchi et al., (2018) regarding SFPQ control of neuronal apoptosis in mice. In this system, SFPQ was more specifically attributed to the regulation of transcription elongation over long introns and its knockout did not result in significant splicing changes. Thus, to establish the specificity for the SFPQ in regulating these retention events, authors would need to show that the same phenotype is not achieved by mis-regulation of other splicing factors. That the authors chose SFPQ based on its binding profile is understandable but potentially confounding given its mechanism of action in transcription of long introns (Takeuchi 2018). Because mechanisms and rates of transcription can influence splicing and exon definition interactions, the role of SFPQ as a transcription elongation factor versus a splicing factor is inadequately disentangled by authors.

To test whether SFPQ acts as a transcription elongation factor, we performed Pol II Cut&Tag in ST2 cells and demonstrated that depletion of SFPQ only caused marginal changes in either the promoter region or gene body of ECM genes, suggesting that the role of SFPQ as a transcriptional activator or elongation factor is minimal (Fig. 7G, 7H). This finding is distinct from SFPQ function in neurons (PMID: 29719248), suggesting that the activation or recruitment of SFPQ in transcriptional regulation may involve tissue-specific factors in neurons.

Reviewer #2 (Public review):

Summary:

The manuscript by Lima et al examines the role of Prmt1 and SFPQ in craniofacial development. Specifically, the authors test the idea that Prmt1 directly methylates specific proteins that results in intron retention in matrix proteins. The protein SFPQ is methylated by Prmt1 and functions downstream to mediate Prmt1 activity. The genes with retained introns activate the NMD pathway to reduce the RNA levels. This paper describes an interesting mechanism for the regulation of RNA levels during development.

Strengths:

The phenotypes support what the authors claim that Prmt1 is involved in craniofacial development and splicing. The use of state-of-the-art sequencing to determine the specific genes that have intron retention and changes in gene expression is a strength.

Weaknesses:

Some of the data seems to contradict the conclusions. And it is unclear how direct the relationships are between Prmt1 and SFPQ.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

First, the claims regarding the effect of PRMT1 loss on splicing are unclear by the section title. In other words, does loss PRMT1 change the incidence of baseline alternative splicing events, or does it introduce new retention events that are responsible for underwriting the craniofacial phenotype? Consistent with this idea, the narrative could benefit from more cellular and/or histological validations of the transcriptomic defects discovered in the RNAseq, which could help contextualize the bioinformatics data with the developmental defects. Moreover, the conclusions drawn about intron retention could be clarified in terms of how applicable the mechanism is likely to be outside of this tissue-specific set of responsive introns.

Loss of Prmt1 did not cause a global shift in intron retention, as shown in Supplemental Figure S2. Instead, Prmt1 deletion caused increase of intron retention specifically in genes enriched in cartilage development, glycosaminoglycan biology, dendrite and axon, and decreased intron retention in mitochondria and metabolism genes (Table. S1). We also tested matrix protein expression by histology to confirm that transcriptomic defects revealed at the RNA level resulted in lower protein production. The new data are in Figure 3E-3H.

Additionally, invoking NMD to align splicing and differential gene expression data understandable but lacking sufficient controls to be conclusive, such as positive control genes to confirm inhibition of NMD.

To validate the blockage of NMD, glutathione peroxidase 1 (Gpx1) intron 1, a well-documented substrate for NMD, is tested as positive control (Fig 4Ac, 4Ad, 9B).

Additionally, it should be clarified whether NMD is a basal mechanism for the regulation of these introns or whether it is an induced mechanism that is invoked by the molecular insult.

In CNCCs, NMD functions both as a physiological mechanism and invoked by molecular insult. Please refer to responses to Reviewer 1’s public review for detailed explanations.

Further, authors present data downstream of two siRNAs for the same gene target, but it remains unclear how siRNAs for the same gene target produce different effects. It may be helpful for authors to clarify how many of the transcriptomic defects are shared versus unique between the siRNAs.

To address this question, we used bioinformatic analysis of the whole genome data to the similarity in changes caused by the two SFPQ-targeting siRNAs. As shown in the new Fig. 7Ba & 7Bb, transcriptomic and intron changes are consistent between the two siRNAs, suggesting that genes targeted by the two siRNA predominantly overlap. This overlap is illustrated by scatter plot analysis of RNAseq DEG and IRI data from each siRNA against SFPQ.

Finally, we stress the importance of presenting the full conceptual basis for SFPQ's potential role in splicing and gene expression. It is significant to note that SFPQ has been previously studied as a splicing factor and was instead determined to function in support of the transcription elongation rather than in splicing. Thus, if authors are confident that the SFPQ manifests directly in splicing changes they encumber the burden of proof to show that its role in transcription, nor another splicing factor, are driving splicing changes.

We demonstrated that depletion of SFPQ only caused marginal changes in either the promoter region or gene body of ECM genes, suggesting that the role of SFPQ as a transcriptional activator or elongation factor is minimal (Fig. 7G, 7H). Please refer to responses to Reviewer 1’s public review for detailed explanations.

Reviewer #2 (Recommendations for the authors):

(1) It is not clear why the authors focused on intron retention targets vs the other possibilities. Skipped Exon is much higher in terms of the number of changes, please clarify. For the intron retention how is this quantified? The traces are nice, but it is hard to tell which part is retained at this magnification. Also, because the focus is on extracellular matrix (ECM) and NMD it would be nice to show some of those targets here. In the tbx1 trace, some are up and some are down. What does that mean for the gene expression?

We have investigated SE initially and found that genes with significant changes in Prmt1 CKO CNCCs fall into diverse functional pathways. Among them, a few genes are critical for skeletal formation, including Postn and Fn, and the function of their exon skipping has been documented. For example, the two exons that are skipped in Postn, Exon17 and 21, have been shown to regulate craniofacial skeleton shape and mandibular condyle hypertrophic zone thickness using transgenic mouse models (PMID: 36859617). As illustrated by Figure 10, the skipped exon of Postn is regulated by multiple splicing factors that may perform overlapping functions in vivo.

Intron retention of each gene is quantified by the ratio of the overall read density of its constitutive intronic regions (CIRs) to the overall read density of its constitutive exonic regions (CERs) and defined as the intron retention index (IRI). In the first section of Response to Reviewer 1’s comments, we explained additional bioinformatic analysis that was performed to address reviewers’ questions, support the confidence of intron event calls and rule out the possibility of other alternative splicing mechanisms, such as by SE, MXE, A5SS or A3SS (Supplemental Figure S5, S6, Table S7).

(2) RNA-Sequencing of Prmt1 mutants nicely shows gene expression changes, including in ECM and GAG genes. While validation of the sequencing results is not necessarily required, it would be very interesting to show the expression in situ. In addition, the heat map shows both downregulated but also upregulated transcripts. This is expected since this protein regulates many genes. However, the volcano plot shows a significant number of genes upregulated. It would be interesting to show what the upregulated genes are. And what is the proposed mechanism for Prmt1 regulation of upregulated genes?

Validation for the transcriptomic changes is shown in Fig. 3E-3H using immunostaining.

As for upregulated genes in Prmt1 mutant, top pathways include cytokine-mediated signaling pathway, signal transduction by p53 signaling pathway and cell morphogenesis (Figure 2E), which are consistent with our previous reports that Prmt1 deletion induces cytokine production in oral epithelium and leads to p53 accumulation in embryonic epicardium (PMID: 32521264, 29420098). Besides these pathways, Prmt1 deletion also caused upregulation of genes involved in adult behavior, postsynaptic organization and apoptotic process, which is consistent with findings from other labs on PRMT1 function in neuronal and cancer cells (PMID: 34619150, 33127433).

(3) Specific transcripts were shown to have elevated intron retention involved in the ECM and GAG pathway. However in Figure 3D it seems to show the opposite with intronic expression decreased and exonic increases and intronic decrease. This is very important to the final conclusion of the paper. In addition, is there a direct relationship between increased intron and downregulation of this specific gene expression? It seems a bit correlational as it could also be an indirect mechanism. One way to test this is to do in vitro translation with and without the specific intron to test if it results in lower expression.

We apologize for the mis-labeling in previous version of Figure 3D, which is now corrected. We also tried to test the direct relationship between intron and downregulation of matrix genes such as Adamts2 using in vitro experiments, however, the introns of matrix genes with high retention tends to be long, many 10 to 50kb in length, making it challenging to generate mini-gene constructs for molecular analysis. We used a different approach and demonstrated that inhibition of NMD with a chemical inhibitor NMDI14 caused dramatic accumulation of the Adamts2, Alpl, Eln, Matn2, Loxl1 and Bgn transcripts, suggesting that retained introns triggered NMD to regulate gene expression and this mechanism acts as a physiological level in CNCCs (Fig. 4). We also blocked NMD in control and Prmt1 null CNCCs, where NMD blockage led to higher accumulation of Adamts2 and Alpl transcripts, suggesting that upon Prmt1 deficiency, NMD is further utilized to degrade intron-containing transcripts (Fig. 4). Similarly, in Sfpq-depleted ST2 cells, blocking NMD caused accumulation of intron-retaining transcripts Col4a2, St6galnac3 and Ptk7 (Fig. 9A, 9B).

(4) While Figure 4 nicely shows the methylation of SFPQ is reduced in Prmt1 CKO cells, it is unclear which reside this methylation occurs. Also the overall expression of SFPQ is also down so it is possible that the methylation is indirect ie Prmt1 regulates some other methyltransferase that regulates SFPQ. Or that because the overall level of SFPQ is down, there is no protein to methylate. How do the authors differentiate between these possibilities?

Previously, arginine methylation of SFPQ has been characterized using in vitro reaction and cell lines with biochemical assays by Snijders., et al in 2015 (PMID: 25605962). Among all PRMTs that catalyze asymmetric arginine dimethylation (ADMA), SFPQ is methylated by only PRMT1 and PRMT3, with PRMT1 showing higher efficiency while PRMT3 showing a lower efficiency. However, PRMT3 is mainly cytosolic. Its expression in CNCCs is about 100-fold lower than PRMT1 (Fig. 1). Based on these knowledges, PRMT1 is the primary arginine methyltransferase for SFPQ, a nuclear protein in CNCCs. We and others have shown in a previous publication that SFPQ methylation on arginine 7 and 9 depends on PRMT1 (PMID: 31451547).

To investigate SFPQ protein degradation in CNCCs, we used MG132 to block proteasomal degradation and observed a partial rescue of SFPQ protein degradation in Prmt1 mutant embryos, suggesting that SFPQ is degraded through proteasomal-mediated mechanism. To address the relationship between SFPQ methylation and protein expression, we assessed arginine methylation of SFPQ that accumulated after MG132 treatment. The accumulated SFPQ was not methylated, confirming the absence of methylation even when SFPQ protein expression is restored.

Snijders., et al, also shown that citrullination induced by PADI4 regulate SFPQ stability (Snijders 2015). We considered this possibility and assessed the expression levels of PADIs. In E13.5 and E15.5 CNCCs, PADI1-4 mRNA expression levels are very low (TPM<5), suggesting that PADIs may not regulate SFPQ stability in CNCCs. A detailed mechanism as to how PRMT1-mediated SFPQ methylation controls stability awaits further investigation.

(5) For the Sfpq deleted experiment, it seems that the two knockdowns are not similar in the gene targets and GO terms different except Wnt signaling. This makes this data difficult to interpret. The genes identified as intron retention are different than the ones identified in Prmt1 deletion and not reduced as much. How does this fit in with the Prmt1 story? If working through Sfpq, it assumes that the targets will be similar and more the 8% would be in common.

To address the first concern, we used bioinformatic analysis of the whole genome data to the similarity in changes caused by the two SFPQ-targeting siRNAs. As shown in the new Fig. 7Ba & 7Bb, transcriptomic and intron changes are consistent between the two siRNAs, suggesting that genes targeted by the two siRNA predominantly overlap. This overlap is illustrated by scatter plot analysis of RNAseq DEG and IRI data from each siRNA against SFPQ.

We have previously identified a group of splicing factors that depends on PRMT1 for arginine methylation, including SFPQ (PMID: 31451547). In the new data in Figures 5, 6 and 10, we tested an additional five PRMT1-dependent splicing factors that are highly expressed in CNCCs: SRSF1, EWSR1, TAF15, TRA2B and G3BP1 (Fig. 5, 6 and 10). Among these factors, SRSF1 and G3BP1 are predominantly expressed in the cytosol of NCCs at E13.5. As splicing activity in the nucleus is needed for pre-mRNA splicing, we excluded these two and focused on the other three proteins. EWSR1 and TRA2B are both methylated in CNCCs and depend on PRMT1 for methylation (Fig. 5). We weren’t able to assess TAF15 methylation because of lack of efficient antibody for the PLA assay. We also demonstrated that their protein expression or subcellular localization was not altered by Prmt1 deletion in CNCCs, unlike SFPQ (Fig. S2). To define their splicing footprint, we performed siRNA-mediated knockdown in ST2 cells, followed by RNA-seq and IRI analysis to define differentially regulated genes and introns, which revealed distinct biological pathways regulated by SFPQ, EWSR1, TRA2B and TAF15, but minimal roles of EWSR1, TRA2B and TAF15 on intron retention when compared to SFPQ (Fig. 10F-10I, Supplemental Figure S7A-S7F). ECM genes are significantly downregulated by all four splicing factors (Fig. 10J-10M), but EWSR1, TRA2B and TAF15 regulate transcription or exon skipping instead of IR, as exemplified by Alpl and Postn (Fig. 10N-10T).

(6) The addition of an NMD mechanism is interesting but not surprising that when inhibiting the pathway broadly, there is an increase in gene expression in the mesoderm cell line. How specific is this to craniofacial development?

NMD is driven by a protein complex composed of SMG and UPF proteins. We show in the revised manuscript that NMD is both a physiological mechanism in CNCCs and triggered by genetic disturbance (Fig. 4). These data are in line with human patient reports where SMG9 mutation in human causes malformation in the face, hand, heart and brain (PMID: 27018474). Mouse genetic studies also demonstrated roles of NMD components during embryonic development.Smg6, Upf1, Upf2 and Upf3a knockout mouse die at early embryonic stages (E5.5-E9.5), and Smg1 gene trap mutant mice die at E12.5 (Han 2018). Additionally, intron retention-triggered NMD in cancer has both promotive and suppressive roles and NMD inhibitors has been tested for cancer therapy and recently cancer immunotherapy. Our findings highlight matrix genes as one of the key targets for NMD during craniofacial development.

Minor:

(1) The supplemental figures are difficult to understand. In the first upload there are many figures and tables, some excel files that are separate uploads and some not. Please upload as separate files so it is clear. And also put them in order that they are in the manuscript.

(2) For the heat map in figure 2B, it would be good to show all the genes or none at all. It seems a bit like cherry-picking to highly only a few. And they are not labeled where they are located in the graph. Are these the top lines if so please label.

(3) Gene names in Figure 3A are difficult to read. I would also not consider BMP7 an ECM gene.

(4) A summary diagram of the interactions proposed will help to make this more understandable.

The supplemental figures are reorganized and uploaded as separate word and excel documents. For Heat map in Fig. 2B, we have removed the gene names. For Fig. 3A, only the most significantly changed gene are labeled in red dots with names. We didn’t label all the genes because of the large number of genes. For the new Figure 3B, we have replaced BMP7. A schematic summary is also added to Supplemental Fig. S9 to illustrate the PRMT1-SFPQ pathway.

Version 2

Abstract

Spliceosomopathies, which are a group of disorders caused by defects in the splicing machinery, frequently affect the craniofacial skeleton and limb, but the molecular mechanism underlying this tissue-specific sensitivity remains unclear. Splicing factors and small nuclear ribonucleoproteins (snRNPs) are core components of splicing machinery, and splicing factors are further controlled by post-translational modifications, among which arginine methylation is one of the most prevalent. We determined the splicing mechanisms in the cranial neural crest cells (CNCCs), a multipotent developmental population that gives rise to the majority of the craniofacial skeleton, and focused on an upstream regulator of splicing proteins, protein arginine methyltransferase 1 (PRMT1). PRMT1 is the highest expressing arginine methyltransferase in CNCCs and its role in craniofacial development is evident from our earlier investigation, where CNCC-specific Prmt1 deletion caused cleft palate and mandibular hypoplasia. PRMT1 catalyzes arginine methylation of splicing factors to modify protein localization, expression and activity. In the present study, we uncover roles of PRMT1 in the regulation of intron retention, a type of alternative splicing where introns are retained in the mature mRNA. CNCCs from mandibular primordium of Prmt1-deficient embryos demonstrated an increase in the percentage of intron-retaining mRNA of matrix genes, which triggered NMD, causing a reduction in matrix mRNA abundance. We further identified SFPQ as a substrate of PRMT1 that depends on PRMT1 for arginine methylation and protein expression in the developing craniofacial structures. Depletion of SFPQ in CNCCs phenocopied PRMT1 deletion whereby matrix, Wnt signaling components and neuronal gene transcripts contained higher IR and exhibited lower expression. We further recognized gene length as a common feature among SFPQ-regulated genes in CNCCs. Altogether, these findings demonstrate that the PRMT1-SFPQ pathway modulates matrix gene expression via IR-triggered NMD in CNCCs during craniofacial development.

Introduction

Craniofacial abnormalities affecting bone formation in the skull and face are the most common birth defects in infants. Proper formation of these structures involves coordination of proliferation, migration, and differentiation of cranial neural crest cells (CNCCs) (Martik & Bronner, 2017; Plein, Fantin, & Ruhrberg, 2015). CNCCs are a transient population of progenitor cells that populate the first and second pharyngeal arches and give rise to the facial skeleton including maxilla, mandible and palates, and the anterior skull (Chai et al., 2000; Jiang, Iseki, Maxson, Sucov, & Morriss-Kay, 2002; Martik & Bronner, 2017). Dysregulation of transcription factors, chromatin remodelers, RNA regulatory proteins, and signaling molecules have been implicated in impaired neural crest development that results in craniofacial defects (Bélanger et al., 2018; Martik & Bronner, 2017; Plein et al., 2015; Strobl-Mazzulla, Marini, & Buzzi, 2012). Spliceosomopathies, which are a group of disorders caused by defects in the splicing machinery, frequently affect the craniofacial skeleton and limb. However, the mechanisms behind the sensitivity of these tissues to spliceosomal defects remain incompletely understood. Splicing factors and small nuclear ribonucleoproteins (snRNPs) are core components of splicing machinery, which control pre-mRNA splicing and mRNA maturation through the process of alternative splicing (AS). AS includes seven types of events, exon skipping, mutually exclusive exons, alternative 5’ and 3’-splice sites, alternative promoters, alternative polyadenylation, and intron retention (Hooper, Jones, Smith, Williams, & Li, 2020). Intron retention (IR), where introns are retained within the mature mRNA sequence, is abundant in plants and fungi (E. T. Wang et al., 2008). Although initially overlooked in mammals due to challenges in analysis, recent technical advances have unveiled its significant roles in various aspects, including cell differentiation and stress response, especially in neuronal, immune, and cancer cells (Braunschweig et al., 2014; Marquez, Brown, Simpson, Barta, & Kalyna, 2012; Monteuuis, Wong, Bailey, Schmitz, & Rasko, 2019; Wong & Schmitz, 2022).

In exploring the splicing mechanisms in CNCCs, we focused on upstream regulators for splicing factors. The activity of splicing factors is tightly controlled by post-translational modifications, among which methylation is the most abundant, surpassing phosphorylation (Ruta, Pagliarini, & Sette, 2021; Thandapani, O’Connor, Bailey, & Richard, 2013). Splicing regulator methylation mostly occurs on arginine motifs, catalyzed by the protein arginine methyltransferase (PRMT) family of enzymes. These methylation events determine their protein stability, subcellular localization, and splicing activity, shaping the splicing product (Blanc & Richard, 2017; Liu & Dreyfuss, 1995; Snijders et al., 2015). PRMT1 is the most abundant PRMT in most cell types. It specifically catalyzes methylation of arginine (R) residues within RG/RGG/GAR repeats, which are motifs enriched in RNA-binding proteins, particularly splicing regulators. This results in asymmetric dimethylation of arginine (ADMA) (Smith, Schurter, Wong-Staal, & David, 2004; Thandapani et al., 2013). The importance of PRMT1 in craniofacial development is evident from our earlier investigation in which genetic deletion of Prmt1 in CNCCs leads to cleft palate and mandibular hypoplasia (Gou, Li, Jackson-Weaver, et al., 2018; Gou, Li, Wu, et al., 2018). In the present study, we uncovered a previously unrecognized role of PRMT1 in regulating intron retention in CNCCs. The mandibular primordium of Prmt1-deficient embryos demonstrated increased retention of introns in mature mRNAs that encode bone and cartilage matrix components. These retained introns contain premature termination codons (PTCs) that trigger nonsense-mediate decay (NMD) to degrade mRNA and reduce gene expression. We further identified SFPQ, EWSR1 and TAF15 as downstream substrates of PRMT1 in the developing craniofacial structures, which depends on PRMT1 for arginine methylation. SFPQ further depends on PRMT1 for protein stability. Depletion of SFPQ in CNCCs partially phenocopied PRMT1 deletion, causing elevated retention of introns in mRNA of matrix, neuronal, and Wnt signaling pathway genes. Matrix and Wnt pathway genes susceptible to perturbation of the PRMT1-SFPQ pathway are further characterized as long genes with a median length of 100kb. Their retained introns trigger NMD to reduce gene expression. Together, these findings demonstrate that the PRMT1-SFPQ pathway modulates CNCC gene expression via control of splicing. Deficiency of this pathway leads to aberrant intron retention, which induces mRNA decay that downregulates matrix expression in CNCCs during craniofacial development.

Results

CNCCs from the embryonic mandibular process display abundant intron retention (IR), which was further elevated by loss of Prmt1

PRMT1 is the most abundant enzyme of the arginine methyltransferase family in cranial neural crest cells (CNCCs), with an expression level ∼10-100 folds higher than other PRMTs (Figure-1A). We previously demonstrated critical roles for PRMT1 in CNCCs during embryogenesis using genetic deletion, whereby neural crest-specific deletion of Prmt1 caused cleft palate and shorter mandibles in mouse (Gou, Li, Jackson-Weaver, et al., 2018; Gou, Li, Wu, et al., 2018). Shorter mandibles, or mandibular hypoplasia is a phenotype shared by most spliceosomopathies, a group of conditions caused by splicing defects (Griffin & Saint-Jeannet, 2020; Merkuri & Fish, 2019). To investigate the molecular mechanisms underlying PRMT1’s role in mandible development and splicing regulation, we conducted a genome-wide analysis of transcriptional and splicing alterations within the embryonic mandibles following Prmt1 deletion. For this purpose, CNCCs, the cell lineage that gives rise to bone and cartilage in the mandible, were isolated from mandibular processes of control (Wnt1-Cre; R26R tdTomato) or Prmt1 deficient (Prmt1 CKO, Wnt1-Cre; Prmt1fl/fl; R26R tdTomato) embryos at embryonic day 13.5 (E13.5) for mRNA extraction and sequencing (Figure-1B, Supplemental Figure-S1). The ensuing comparative analysis using rMATS (Y. Wang et al., 2024) unveiled changes across five types of alternative splicing events (Figure-1C), echoing our earlier finding where PRMT1 regulates exon usage in epicardial cells during cardiac morphogenesis (Jackson-Weaver et al., 2020). In contrast to embryonic epicardial cells, we noted a higher abundance of IR changes in mandibular CNCCs (Figure-1C). The IR changes were exemplified by the increase of intron retention in Pex12, Mmp23, and Ecm1 (Figure-1D, 1F), and the decrease of retained intronic expression in Tbx1 (Figure-1E, 1F). To comprehensively analyze intron retention at a gene-specific level, we employed IRI (intron reads index), which is an algorithm that quantifies the ratio of normalized read counts between intronic and exonic regions (Ni et al., 2016; Sun et al., 2023; Tian et al., 2020). IRI analysis demonstrated that Prmt1 deficiency didn’t cause a global shift in intron retention levels (Supplemental Figure-S2) but changed intron retention in genes of multiple biological pathways (Supplemental Table S1). We noted protein binding, metal ion binding, and cartilage development among genes that present increased intron retention, while mitochondrion and metabolic pathways were identified among genes with decreased intron retention (Supplemental Table S1). These findings identify intron retention as a prevalent phenomenon in CNCCs during craniofacial development, and the deletion of Prmt1 further enhanced IR in genes involved in diverse biological processes.

CNCC-specific deletion of Prmt1 elevates intron retention in the embryonic mandibular process.

(A) Expression levels of PRMT1-9 mRNAs in primary isolated cranial neural crest cells (CNCC) from Wnt1-Cre; R26RtdTomato mouse embryo heads at E13.5 and E15.5. TPM, transcript per million. (B) Diagram illustrating the isolation of CNCCs from embryonic mandibles, followed by poly(A)+ mRNA isolation and sequencing (n=4 in control and Prmt1 CKO group). (C) Prmt1 deletion in CNCC caused changes in alternative splicing (AS). Changes in AS events were analyzed by rMATS using RNA sequencing data and significant changes in each type of AS were shown in a stacked bar chart. SE, skipped exon. IR, intron retention. MXE, mutually exclusive exons. A5SS, alternative 5’ splice site. A3SS, alternative 3’ splice site. (D & E) Intron retention was prevalent in CNCCs and altered by Prmt1 deletion. Track view of genes demonstrating intronic and exonic expression with blue boxes indicating exons and blue lines indicating introns. Intron retention was elevated by Prmt1 deletion in Pex12, Mmp23, and Ecm1 (D) and reduced in Tbx1 (E). (F) Quantification of intron expression in Pex12, Mmp23, Ecm1 and Tbx1 by the percentage of intron-retaining mRNAs in each gene, calculated from IRI analysis based on RNA-seq data. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; R26RtdTomato. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; R26RtdTomato.

Neural crest-specific deletion of Prmt1 caused a significant reduction of matrix gene expression in the developing mandibles

The main documented role for intron retention is to reduce expression of the gene harboring retained introns via non-sense mediated decay (NMD), the cytosolic RNA surveillance system (Schmitz et al., 2017). When intron-retaining mRNA transcripts are loaded onto the translation machinery, introns that contain premature termination codons (PTCs) within their open reading frames trigger NMD, leading to mRNA transcript degradation. Reduction in mRNA abundance leads to decreased protein expression for genes harboring these introns (Wong, Au, Ritchie, & Rasko, 2016). The intron-retaining transcripts illustrated in Figure-1, Pex12, Mmp23, and Ecm1 demonstrated increased intronic expression and decreased exonic expression, supporting this negative correlation (Figure-1D, 1F). To systematically analyze genes that are differentially downregulated, we compared the transcription profile between control and Prmt1 CKO mandibular CNCCs and revealed downregulation of 303 genes and concurrent upregulation of 160 genes by Prmt1 deletion (Figure-2A, 2B). Upon pathway analysis, glycosaminoglycan (GAG) degradation (35 out of 303 genes) and extracellular matrix (ECM) (33 out of 303 genes) emerged as top pathways among downregulated genes (Figure-2C). These GAG degradation and ECM genes encode matrix proteins pivotal to osteogenic/chondrogenic differentiation and the formation of bone and cartilage matrix. Ingenuity Pathway Analysis (IPA) of downregulated genes further suggested a disruption in bone, cartilage, and connective tissue development within Prmt1 CKO mutant mandibles (Figure-2D). In addition, Prmt1 deficiency upregulated genes in cytokine-mediated signaling pathway and p53 signaling, which are consistent with previous reports that Prmt1 deletion resulted in cytokine production in oral epithelium and p53 accumulation in embryonic epicardium (Jackson-Weaver et al., 2020; Zhang et al., 2018). Prmt1 deletion also upregulated genes involved in adult behavior, postsynaptic membrane organization and regulated exocytosis, which echoes findings on PRMT1 function in the neuronal lineages and cancer (Chen et al., 2021; Hashimoto, Fukamizu, Nakagawa, & Kizuka, 2021; Hashimoto, Kumabe, et al., 2021; Hashimoto et al., 2022; Xu & Richard, 2021) (Figure-2E).

CNCC-specific Prmt1 deletion reduced matrix gene expression in the developing mandibles.

(A) Volcano plot illustrating upregulation of 160 and downregulation of 303 genes in the mandibular primordium of Prmt1-deficient embryos at E13.5, compared to control mandibles. (B) Heatmap showing differential gene expression between control and Prmt1-deficient mandibles. (C) GO analysis of pathway enrichment in downregulated genes demonstrating glycosaminoglycan (GAG) degradation and extracellular matrix (ECM) organization as the top pathways. (D) Ingenuity Pathway Analysis suggesting connective tissue, bone, and cartilage development as affected biological processes based on downregulated genes. (E) GO analysis of pathway enrichment in upregulated genes demonstrating adult behavior and cytokine-mediated signaling and p53 signal transduction as top pathways. Control: Wnt1-Cre; R26RtdTomato (n=4). Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; R26RtdTomato (n=4).

Deletion of Prmt1 in CNCCs elevated intron retention in mRNAs that encode ECM and GAG degradation enzymes

To determine whether downregulation of matrix genes in Prmt1 CKO embryos is regulated by IR-mediated mechanism, we first focused on ECM genes downregulated in Prmt1-deficient mandibles and investigated their intron retention events. The findings are presented in a scatter plot where the red line denotes unchanged IR. A prominent elevation of IR within ECM genes was revealed by the fact that most ECM data points landed above the red line (Figure-3A). ECM genes with the most significant IR increase were labeled and highlighted in red. The IR increase was also exemplified by Adamts16 and Cthrc1 (Figure-3B, 3C), where track view of the two genes illustrated increased IR and reduced exonic expression (Figure-3Ba, 3Bb). To validate the reproducibility of IR elevation in ECM transcripts, we collected four additional embryos from each group, isolated mature mRNA using poly(A)+ magnetic beads, and designed primers for intronic and exonic region of additional ECM genes with RT-PCR analysis. In line with our initial observations, intronic expression increased significantly, coupled with decreased exonic expression in Dcn, Tnn, Lox, Loxl1, Matn2, Col14a1, Adamts12, Adamts2, Adam12, and Scara5 of the Prmt1 CKO CNCCs as compared to control (Figure-3D). The decrease in exonic expression, representing mRNA abundance, led to a decline in protein expression, as demonstrated by immunostaining for LOXL1 and FBLN5 in CNCCs (Figure-3E-H). These findings demonstrate that ECM transcripts in Prmt1 CKO mandibles exhibited IR elevation and reduction in mRNA expression, leading to lower ECM protein production. Next, we analyzed the intron retention events in GAG degradation genes, which is the top downregulated gene cluster in the Prmt1 CKO group and clinically associated with craniofacial birth defects (Mizumoto & Yamada, 2021; Paganini, Gramegna Tota, Superti-Furga, & Rossi, 2020; Schwartz & Domowicz, 2002). Similar to ECM genes, GAG degradation genes showed IR elevation in scatter plot where most data points landed above the red line (Figure-3I). The track view of St6galnac3 and Galn11 further illustrated the increased IR and lower mRNA abundance (Figure-3J, 3K). Taken together, ECM and GAG degradation genes that are downregulated in Prmt1 CKO embryos demonstrate elevated intron retention.

PRMT1 regulates intron retention in ECM and GAG degradation genes.

(A-C) Intron retention increased in the majority of ECM gene transcripts that were downregulated in Prmt1 deficient embryos, as illustrated by a scatter plot based on intron retention index (IRI). The red line delineates unchanged levels of intron retention. Genes with the top differential IR were represented by red dots and labeled. ECM gene transcripts Adamts16 and Cthrc1 demonstrating higher intron retention in Prmt1- deleted embryos were illustrated by track view in B and quantified for intronic (left) and exonic (right) expression as shown in C. (D) Higher intron retention and lower mRNA abundance of ECM genes were validated in additional embryo samples by RT-PCR. Primers that span the intronic or intron-exon junction region were used to assess intronic expression. Primers that span the exonic region were used to examine exonic expression that indicated mRNA abundance. (E-H) Reduced expression of LOXL1 and FBLN5 was examined at the protein level by immunostaining (E, G) and quantified (F, H). Ee and Ef illustrated the plane of section and the region of analysis for E and G. (I-K) Intron retention increased in the majority of GAG degradation gene transcripts that were downregulated in Prmt1 deficiency, as indicated by a scatter plot based on IRI. The red line defines where intron retention is unchanged. Genes were represented by black dots and labeled in red. GAG degradation genes St6galnac3 and Galnt11 demonstrating higher intron retention in Prmt1-deleted embryos were illustrated by track view (J) and quantified for intronic (left) and exonic (right) expression (K). * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; R26RtdTomato. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; R26RtdTomato. Scale bar: (E,G), 100µm.

IR-triggered NMD acts as a physiological and stress-induced mechanism for mRNA decay in CNCCs

We inspected the retained introns in these ECM and GAG degradation transcripts and identified PTCs within their open reading frames, suggesting PTC-triggered NMD as a potential mechanism for transcript decay (Supplemental Table S2) (Wong et al., 2016). To test this hypothesis, we inhibited the NMD machinery using a chemical inhibitor NMDI14, which disrupts the interaction between UPF1 and SMG7 to block NMD complex formation and activity (Martin et al., 2014). To this end, Wnt1-Cre; Prmt1fl/+; R26R tdTomato were bred with Prmt1fl/fl; R26R tdTomato mice to generate Prmt1 CKO (Wnt1-Cre; Prmt1fl/fl; R26R tdTomato) and Prmt1 heterozygous embryos as littermate control (Prmt1 Het, Wnt1-Cre; Prmt1fl/+; R26R tdTomato). Primary CNCCs from E13.5 Prmt1 CKO and Prmt1 Het were plated and treated with NMDI14 or DMSO as control, followed by poly(A)+ mRNA extraction and RT-PCR analysis using primers that span the intronic or exonic regions. We first validated the inhibition of NMD activity by NMDI14 through analysis of glutathione peroxidase 1 (Gpx1) intron 1, which is a well-documented substrate for NMD-mediated transcript decay (Sun, Moriarty, & Maquat, 2000). We further demonstrated that NMDI14 caused a significant increase of Adamts2 intron 19 and Alpl intron 5 expression in the control CNCCs from Prmt1 Het (Figure-4Aa), suggesting that inhibition of NMD blocked IR-triggered mRNA decay, thereby causing accumulation of introns. A more prominent accumulation of introns was induced by NMDI14 in the Prmt1 CKO CNCCs, accompanied by significant increase of mRNA abundance indicated by higher exonic expression levels (Figure-4Ab). These data suggest that Prmt1 deficiency invoked IR-induced NMD for mRNA decay which reduced mRNA abundance.

IR-triggered NMD functions as a basal and stress-responsive mechanism for mRNA decay in CNCCs.

(Aa, Ab) Treatment with the NMD inhibitor NMDI14 (NMDI) led to accumulation of intron-retaining (intron) transcripts of Gpx1, Adamts2, and Alpl in CNCCs from Control (Prmt1 Het) and Prmt1 CKO embryos, analyzed by RT-PCR. Gpx1 serves as a positive control to validate NMD inhibition by NMDI14. (Ba–Bf) NMDI14 caused accumulation of intron-retaining (intron) and total mRNAs (exon) of Adamts2, Alpl, Eln, Matn2, Loxl1 and Bgn in CNCCs. CNCCs were isolated from E13.5 Wnt1-Cre; R26RtdTomato and analyzed by RT-PCR. Primers that span the intronic or intron-exon junction region were used to assess intronic expression. Primers that span the exonic region were used to examine exonic expression that indicated mRNA abundance. (Ca, Cd) Intron retention of matrix transcripts Adamts2 and Fbln5 was detected in four independent control embryos by RNAseq, as illustrated by track views. (Cb-Cf) NMDI14 treatment caused accumulation of intron-retaining Adamts2 and Fbln5 transcripts. CNCCs isolated from E13.5 Wnt1-Cre; R26RtdTomato embryos were treated by DMSO or NMDI14 (NMDI), followed by mRNA extraction and assessment with semi-quantitative PCR. Primers (indicated by the red arrows) were designed to span regions (red line) of intron 19 of Adamts2 or intron 7 of Fbln5. * p<0.05 NMDI vs. DMSO of the same group.

We noted significant accumulation of retained introns in control CNCCs from Prmt1 Het (Figure-4Aa), suggesting a possibility that IR-induced NMD is a basal or physiological mechanism to regulate mRNA levels in CNCCs. To assess this possibility, we isolated CNCCs from control Wnt1-Cre; R26R tdTomato embryos at E13.5 and treated cells with NMDI14 or DMSO, followed by poly(A)+ mRNA extraction and RT-PCR analysis. Multiple ECM genes were assessed for the role of NMD, which revealed significant accumulation of retained introns within Adamts2, Alpl, Eln, Matn2, Loxl1 and Bgn, accompanied by increased mRNA abundance following NMDI14 treatment (Figure-4Ba-Bf). To strengthen these findings, we used traditional semi-quantitative PCR to detect a longer span (600-1000bp) of the intronic region. In the NMDI14-treated group, accumulation of Adamts2 intron 19 and Fbln5 intron 7 were revealed by higher intensity bands compared to DMSO group (Figure-4Ca-4Cf). Overall, these findings demonstrate that the NMD process is responsible for the decay of ECM transcripts. More importantly, we defined IR-triggered NMD as a physiological mechanism in CNCCs to control ECM mRNA abundance and gene expression during craniofacial development. Upon molecular stress caused by Prmt1 deletion, NMD is further exploited to disrupt ECM expression.

PRMT1 methylates splicing factors SFPQ, EWSR1 and TAF15 in CNCCs

Next, we investigated the molecular mechanism by which Prmt1 deletion disrupts intron retention. PRMT1 is a methyltransferase that catalyzes methylation of arginine (R) residues within RG/RGG/GAR repeats. These repeats are highly enriched in RNA-binding proteins, particularly splicing regulators (Thandapani et al., 2013). PRMT1 generates asymmetric dimethylation (ADMA) on arginine to regulate the turnover, subcellular localization, and activity of splicing regulators (Smith et al., 2004). We therefore hypothesized that PRMT1 regulates the retention of introns via methylation of splicing regulators. Earlier studies by Graham et al. and our team have characterized the landscape of arginine methylation governed by PRMT1 and identified a cohort of RNA processing proteins exhibiting PRMT1-dependent methylation (Hartel, Chew, Qin, Xu, & Graham, 2019). Among these, six splicing regulators are highly expressed in embryonic CNCCs: SFPQ, SRSF1, EWSR1, TAF15, TRA2B, and G3BP1 (Figure-5A). Given that pre-mRNA splicing occurs in the nucleus following transcription, we examined the subcellular localization of these splicing factors in control CNCCs from E13.5 mouse embryos. SFPQ, EWSR1 and TAF15 predominantly localized to the nucleus (Figure-5Ba-5Bl). TRA2B was distributed in both nucleus and cytoplasm (Figure-5Bm-p). While SRSF1 and G3BP1 were primarily expressed in the cytoplasm (Figure-5Bq-x). These observations were corroborated by quantitative analysis of the nuclear to cytoplasmic ratio of their expression (Figure-5C), suggesting nuclear proteins SFPQ, EWSR1, TAF15 and TRA2B as prime candidates to mediate PRMT1-controlled splicing activity. All four splicing factors are documented substrates for PRMT1-catalyzed methylation in Hela, Jeg3 or HEK293 epithelial cells (Hartel et al., 2019; Jobert, Argentini, & Tora, 2009; K. K. C. Li, Chau, & Lee, 2018; Pahlich, Bschir, Chiavi, Belyanskaya, & Gehring, 2005; Snijders et al., 2015). To determine whether PRMT1 is responsible for methylation of these splicing factors in the craniofacial structures, we employed proximity ligation assay (PLA), a technique designed for in-situ detection of protein modifications. To this end, E13.5 embryo sections were probed with anti-SFPQ antibody and anti-ADMA antibody that detects methyl-arginine using the PLA kit, which revealed the presence of methylated SFPQ as distinct green punctate staining. Within the craniofacial complex, SFPQ methylation was observed in the mandibular processes at E13.5 (Figure-5Da-5Dc, 5E). In Prmt1 CKO embryos, the signal of methyl-SFPQ markedly decreased within the CNCC-derived mesenchyme region (Figure-5Dd-5Df, 5E). The same analysis was conducted in the maxillary regions, which also demonstrated robust SFPQ methylation in the control group and a dramatic reduction in Prmt1-deleted embryos (Figure-5Fa-5Ff, 5G). This CNCC-specific deletion of PRMT1 only reduced SFPQ methylation in the CNCCs, as non-CNCC lineages within the epithelium continued to display methyl-SFPQ signals (Supplemental Figure-S3A). To detect EWSR1 methylation, E13.5 embryo sections were probed with anti-EWSR1 antibody and anti-ADMA antibody using PLA kit. A low level of EWSR1 methylation was observed in both mandibular and maxilla processes at E13.5 (Figure-5H, 5I). Using the same approach, we also detected TRA2B methylation in mandibular and maxilla processes at E13.5 (Figure-5J, 5K), but the levels of EWSR1 and TRA2B methylation were much lower compared to SFPQ. In contrast to the low levels of methylation in craniofacial structures, the embryonic abdominal tissue showed prominent methylation of EWSR1 and TRA2B, validating the PLA techniques and EWSR1/TRA2B antibodies in efficient detection of their methylation (Supplemental Figure-S3B, S3C). We further determined whether methylation of EWSR1 and TRA2B depends on PRMT1 by comparing methylation signal between control and Prmt1 deficient embryos and detected a significant reduction of methyl-EWSR1 and methyl-TRA2B within the CNCC-derived mesenchyme region in Prmt1 CKO samples (Figure-5I-5K). TAF15 methylation wasn’t assessed due to lack of a compatible antibody. Collectively, these data indicate that SFPQ, EWSR1 and TRA2B are methylated by PRMT1 in the CNCCs at E13.5 within the craniofacial complex, with SFPQ showing the most prominent methylation and in a PRMT1-depedent manner.

PRMT1 methylates SFPQ, EWSR1 and TRA2B in CNCCs.

(A) Expression levels of splicing factors SFPQ, SRSF1, EWSR1, TAF15, TRA2B, HnRNPA1, WDR70 and G3BP1 in primary isolated CNCCs from Wnt1-Cre; R26RtdTomato mouse embryonic heads at E13.5 and E15.5. TPM, transcript per million. (B-C) Subcellular localization of splicing factors in CNCCs by immunostaining using sagittal sections of the mandibular process from wild-type mouse embryos, which revealed nuclear expression of SFPQ (a-d), EWSR1 (e-h), and TAF15 (i-l); even distribution of TRA2B between nucleus and cytosolic compartment (m-p), and cytoplasmic expression of SRSF1 (q-t) and G3BP1 (u-x). The subcellular distribution was quantified in C and presented as nuclear to cytosolic signal ratio (Nuc/Cyto Ratio). (D-G) SFPQ methylation diminished in the mandibular (D&E) and maxillary (F&G) processes of Prmt1 deficient embryos at E13.5. (H-K) Reduction of EWSR1 and TRA2B methylation was observed in the mandibular processes of Prmt1 deficient embryos at E13.5. Methylation was detected by proximity ligation assay (PLA). Green puncta indicated methyl-SFPQ, TRA2B or EWSR1. Nuclei were counterstained with DAPI (blue). Representative images are shown for Control (Da-Dc; Fa-Fc; Ha-Hc; Ja-Jc) and Prmt1 CKO (Dd-Df; Fd-Ff; Hd-Hf; Jd-Jf). Higher magnification views in (Dc, Df, Fc, Ff, Hc, Hf, Jc, Jf) illustrate methyl-SFPQ, TRA2B and EWSR1 (green puncta) in the nuclei, as indicated by white arrows. E, G, I and K showed quantification of PLA puncta normalized to cell number in four biological replicates, presented as mean ± SEM. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; R26RtdTomato. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; R26RtdTomato. Scale bar = 100µm in B, D, F, H and J except in enlarged panels, where scale bar = 25µm (Bd, Bh, Bl, Bp, Bt, Bx, Dc, Df, Fc, Ff, Hc, Hf, Jc, Jf).

PRMT1-catalyzed methylation protects SFPQ from proteasomal degradation

PRMT1-induced methylation controls the activity of splicing factor by modifying their stability, subcellular distribution, or RNA binding affinity (Rho, Choi, Jung, & Im, 2007). We first focused on SFPQ and examined whether SFPQ protein expression or subcellular localization was disturbed by Prmt1 deletion. In control embryos, SFPQ exhibited prominent expression in CNCCs and mainly localized in the nucleus within the mandibular and maxillary processes (Figure-6Aa-c, 6Ca-c, 5Ba-d, 5C). In Prmt1-deficient CNCCs, SFPQ protein expression level declined significantly in the maxilla and mandible (Figure-6A-D). A corroborating Western blot (WB) analysis using tissues from the craniofacial structures, including facial and anterior skull regions of embryonic heads, confirmed this decline in SFPQ expression (Figure-6E, 6F). These data suggest two possible mechanisms: PRMT1 promotes SFPQ expression or protects SFPQ from degradation. We first analyzed SFPQ mRNA levels in control and Prmt1-deficient CNCCs, which showed no discernable difference (Figure-6G), suggesting that loss of PRMT1 didn’t reduce SFPQ mRNA expression. Next, we tested whether PRMT1 regulates SFPQ degradation and assessed the proteosome-mediated degradation by treating CNCCs with MG132, a proteasome inhibitor. MG132 treatment in control CNCCs from Prmt1 Het embryos did not alter the expression of SFPQ, but in the Prmt1 CKO CNCCs, MG132 caused a significant restoration of SFPQ protein (Figure-6H, 6I), suggesting that SFPQ protein in Prmt1-deficient CNCCs was reduced by proteasomal degradation. To test the relationship between arginine methylation and SFPQ degradation, we assessed SFPQ methylation in these MG132-treated CNCCs. SFPQ proteins rescued by MG132 in Prmt1-deficient CNCCs was not methylated (Figure-6J, 6K), suggesting that PRMT1-catalyzed arginine methylation may protect SFPQ from proteasomal degradation in CNCCs.

PRMT1 depletion reduces SFPQ protein levels via proteasomal degradation.

(A-D) SFPQ protein level was significantly reduced in the mandibular (A&B) and maxillary (C&D) processes of Prmt1 deficient embryos. SFPQ protein was detected by immunostaining and quantified in B and D. (E & F) SFPQ protein levels declined dramatically in the Prmt1 deficient embryonic head, as detected by Western blotting (E) and quantified with ImageJ (F). (G) PRMT1 deletion did not alter SFPQ mRNA levels in CNCC. (H-K) SFPQ protein accumulated in Prmt1-deficient CNCCs upon MG132 treatment, and the accumulated SFPQ protein remained un-methylated. CNCCs isolated from Prmt1 CKO embryos and littermate controls (Prmt1 heterozygous) were treated with MG132. SFPQ protein was detected by immunostaining in H and quantified in I. SFPQ methylation was detected by PLA in J and quantified in K.* p<0.05 Prmt1 CKO vs Control. Control, Wnt1-Cre; R26RtdTomato. Prmt1 CKO, Wnt1-Cre; Prmt1fl/fl; R26RtdTomato. Scale bar =100µm in A & C. Scale bar =25µm in H & J.

We further examine whether protein expression of EWSR1, TAF15 and TRA2B was disturbed by Prmt1 deletion in the craniofacial complex. Unlike SFPQ, EWSR1, TAF15 and TRA2B showed a similar level of protein expression in Prmt1 CKO and control embryos (Supplemental Figure-S4A-S4F). These findings were supported by WB analysis using tissues from the craniofacial structures (Supplemental Figure-S4G).

The subcellular distribution of EWSR1, TAF15 and TRA2B wasn’t altered either (Supplemental Figure-S4H-S4J). In summary, loss of Prmt1 reduced SFPQ protein expression via proteasomal degradation, whereas the protein expression and subcellular localization of EWSR1, TAF15 and TRA2B were not altered by Prmt1 deficiency.

SFPQ regulates the splicing of matrix, Wnt signaling and neuronal genes in CNCCs

We further investigated the role of SFPQ in CNCCs. Primary CNCCs were isolated from control (Wnt1-Cre; R26RtdTomato) embryos at E13.5 and enriched by Td+ signal using cell sorting. Within a day, Td+ CNCCs were transfected with control siRNA or two independent SFPQ-targeting siRNAs to deplete SFPQ, and then poly(A)+ mRNA was purified followed by sequencing (Figure-7A). To determine whether these primary isolated CNCCs retained neural crest cell characteristics after siRNA-mediated knockdown, the expression of CNCC markers Twist1, Sox10, Msx1, Snai2, and Tfap2a was examined (Achilleos & Trainor, 2012; Ishii et al., 2012). The CNCC marker expression in transfected CNCCs after isolation is comparable to fresh CNCCs from E13.5 or E15.5 embryos (Supplemental Figure-S5), validating CNCC identity and supporting this method as a feasible approach for mechanistic study of neural crest cell biology using developmental stage-specific embryos. Subsequently, using sequencing data from these siRNA-transfected CNCCs which exhibit about 50% reduction of Sfpq (Figure-7A), we conducted bioinformatic analyses of differentially expressed genes (DEG) and differentially retained introns (IRI) between control and Sfpq-depleted CNCCs. First, we confirmed that the transcriptomic and intronic changes induced the two independent siRNAs targeting SFPQ are predominantly overlapping, as illustrated by scatter plot analysis of the DEG and IRI data (Figure-7Ba, 7Bb). Next, we focused on changes in retained introns caused by Sfpq depletion and revealed higher intron retention in 397 (siSFPQ #1 vs. siCont) and 388 (siSFPQ #2 vs. siCont) genes and lower IR in 129 (siSFPQ #1 vs. siCont) and 151 (siSFPQ #2 vs. siCont) genes (Figure-7Ca, Cb). Genes with higher IR also bear significant overlap with downregulated genes. Among 269 (siSFPQ #1 vs. siCont) and 232 (siSFPQ #2 vs. siCont) genes downregulated by Sfpq depletion, 81 (siSFPQ #1 vs. siCont) and 60 (siSFPQ #2 vs. siCont) genes overlapped with elevated IR, respectively (Figure-7Cc, Cd). Sfpq depletion also led to downregulation of 269 (siSFPQ #1 vs. siCont) and 232 (siSFPQ #2 vs. siCont) genes, among which 81 (siSFPQ #1 vs. siCont) and 60 (siSFPQ #2 vs. siCont) genes overlapped with elevated IR, respectively (Figure-7Cc, Cd). Genes showing higher intron retention in Sfpq depleted CNCCs encompassed “proteoglycan metabolic process” and “proteoglycan biosynthesis process”, which included GAG degradation genes, and “connective tissue development”, which includes ECM genes (Figure-7Da, 7Db). We also analyzed genes that were differentially expressed in control vs. Sfpq-depleted CNCCs (Figure-7E). GO analysis of differentially downregulated genes by SFPQ-specific siRNAs further revealed extracellular matrix and extracellular structure as the top pathways suppressed by SFPQ depletion, suggesting that IR-mediated regulation of ECM is a key functional consequence following Sfpq deficiency (Figure-7F). SFPQ has demonstrated direct roles in transcriptional regulation, where it facilitates RNA polymerase II (Pol II) recruitment and transcriptional termination in neuronal lineages (Takeuchi et al., 2018). To determine whether SFPQ directly regulates the transcription of matrix genes, we performed CUT&TAG with antibody again Rbp1, the largest subunit of Pol II, in ST2 cells, which is a mesenchymal cell line used for mechanistic studies of osteogenic differentiation (Canales et al., 2023; Ishida, Kawao, Mizukami, Takafuji, & Kaji, 2021; K. Li et al., 2019; Mizukami et al., 2023; Pregizer, Barski, Gersbach, García, & Frenkel, 2007; Seong et al., 2023; Strauss et al., 2019; Tu et al., 2007; Yang et al., 2019). ST2 cells were transfected with control or SFPQ-targeting siRNAs and collected for CUT&TAG followed by analyses of Pol II binding at the promoter region (3kb ± TSS) and gene body of matrix genes. Pol II recruitment was similar between control and SFPQ-depleted groups with marginal differences (Figure-7G, 7H), suggesting that transcriptional repression cannot fully account for the downregulation of matrix gene in SFPQ depleted groups.

PRMT1-SFPQ pathway regulates matrix genes in CNCCs.

(A) SFPQ knockdown caused around 50% reduction in Sfpq expression in CNCCs. CNCCs were transfected with two independent siRNAs targeting SFPQ, or control siRNA, followed by poly(A)+ mRNA extraction and RNA sequencing. * p<0.01 siSFPQ vs siControl. (B) The two independent siRNAs targeting SFPQ caused similar transcriptomic and intronic changes, shown by scatter plot of DEG (Ba) and IRI (Bb). DEG, differentially expressed genes. IRI, intron retention index. (C) SFPQ depletion altered intron retention in CNCCs. Changes of IR events in genes showing increased (Up, yellow color) or decreased (Down, orange color) IR were illustrated by stacked bar graphs. Pie chart demonstrated differentially regulated genes, with shaded areas among downregulated genes (orange) highlighting their overlap with IR elevated genes (red). (D) GO analysis of genes with elevated IR following SFPQ depletion. (E&F) Heatmap and GO analysis of SFPQ-regulated genes in CNCC. (G&H) Pol II CUT&Tag analysis in ST2 cells transfected with control or SFPQ siRNAs showing Pol II recruitment in downregulated and upregulated genes (G), promoter regions (Ha) and gene body (Hb). P-value was indicated at the top. siControl, control siRNA. siSFPQ#1 and siSFPQ#2, two independent SFPQ siRNAs.

More intriguingly, Wnt signaling emerged as the top pathway regulated by SFPQ in CNCCs (Figure-Figure-7Da, 7Db). Sfpq depletion increased the percentage of intron-retaining mRNAs of genes in this cluster and reduced total mRNA expression (Figure-8A). The Wnt signaling components-regulated by SFPQ encode both positive and negative regulators of Wnt signaling and encompass both canonical and non-canonical Wnt signal regulators. Additionally, neuronal genes constitute a big fraction among SFPQ-regulated introns as “axon guidance”, “axonogenesis”, “neuron projection development”, and “leaning or memory” emerged as the top biological processes among differentially upregulated introns and SFPQ-regulated genes, and “regulation of membrane potential”, “learning or memory” “cognition” and “locomotory behavior” denotes differentially downregulated genes (Figure-7F). In CNCCs, Sfpq depletion increased the percentage of intron-retaining mRNAs of these neuronal genes and reduced total mRNA expression (Figure-8B). Altogether, these findings demonstrate that depletion of SFPQ enhanced intron retention and reduced mRNA expression in matrix, Wnt signaling components and neuronal genes.

SFPQ regulates intron retention of Wnt signaling and neuronal genes in CNCCs.

SFPQ depletion in CNCCs elevated intron retention and decreased mRNA abundance of Wnt signaling components (A) and neuronal genes (B). The levels of mRNA abundance (left) and intron retention (right) were illustrated by a two-sided bar graph. TPM, transcripts per million. * p<0.01 siSFPQ vs siControl. siControl, control siRNA. siSFPQ#1 and siSFPQ#2, two independent SFPQ siRNAs.

SFPQ depletion reduces matrix and Wnt signaling gene expression via IR-induced NMD

In Sfpq-depleted CNCCs, retained introns also contained PTCs within their open reading frames, suggesting the involvement of PTC-triggered NMD (Supplemental Table S3). To investigate the role of IR-induced NMD, we utilized ST2 cells. SFPQ knockdown in ST2 cells caused an increase in retained introns and decrease in mRNA expression for Col4a2, St6galnac3, and Ptk7 (Figure-9A). To determine whether Col4a2, St6galnac3, and Ptk7 mRNA were degraded through IR-induced NMD, we inhibited NMD with NMDI14 (Martin et al., 2014). NMDI14 treatment caused accumulation of intron 4 of Col4a2, intron 1 of St6galnac3, and intron 1 of Ptk7 (Figure-9B), indicating that the decay of intron-retaining transcripts in the Sfpq-depleted cells depends on the NMD machinery. To assess the impact of NMD on the abundance of these mRNAs, we designed PCR primers to detect the exonic regions and observed that inhibition of NMD restored and increased the abundance of these mRNAs in Sfpq-depleted ST2 cells (Figure-9C). These data indicate that depletion of Sfpq causes aberrant intron retention that triggers mRNA decay through NMD. We further noted that NMDI14 induced accumulation of intron 4 of Col4a2 and intron 1 of St6galnac3 in the control group (Figure-9B), accompanied by accumulation of their transcripts (Figure-9C), suggesting that NMD is also a basal mechanism to regulate gene expression in ST2 cells, a function similar as in embryonic CNCCs.

SFPQ depletion reduces long gene expression via intron retention triggered NMD.

(A) SFPQ depletion promoted intron retention and reduced mRNA abundance of Col4a2, St6galnac3, and Ptk7 in ST2 cells. Bar chart showing RT-PCR analysis of intronic and exonic expression. (B&C) NMD inhibitor NMDI14 caused the accumulation of retained introns and total mRNAs of Col4a2, St6galnac3, and Ptk7 in ST2 cells. Bar chart showing RT-PCR analysis of intronic (B) and exonic expression (C) in DMSO or NMDI-treated cells. NMDL14-mediated inhibition of NMD was validated using Gpx1 as a positive control. *p<0.05 siSFPQ vs. siControl. # p<0.05 NMDI vs. DMSO treatment of the same group. (D) SFPQ binding peaks were mapped to retained intron 1 but not spliced intron 6 of Ptk7. Peak distribution from published Sfpq CLIP-seq data using E13.5 brain (top) and track view of RNA-seq data using siControl or siSFPQ-transfected CNCCs (bottom) for Ptk7, with the retained Ptk7 intron 1 (red box) and spliced intron 6 (green box) highlighted. (E) SFPQ binding peaks were significantly enriched in retained intron regions within CNCCs. Violin plot displaying the density of SFPQ binding peaks in introns with elevated retention compared to introns with reduced retention or no change. P-value was calculated using Mann-Whitney U test and indicated at the top. (F) SFPQ binding peaks were preferentially enriched in genes with higher intron retention when compared to genes with no IR change. P = 0.07 using Fisher’s exact test. (G) SFPQ-regulated genes were significantly longer than average. Violin plot displaying the distribution of length for genes showing increased intron retention (IR Up), decreased intron retention (IR Down), or unchanged intron retention (No Change). Median length of the IR Up group was highlighted with solid blue line. Median length of the IR Down and No Change groups was highlighted with dotted blue line. P-value was calculated using Mann-Whitney U test and indicated at the top.

Disturbance of SFPQ activity impaired splicing of long genes and long introns

SFPQ predominantly binds to intronic regions and regulates splicing (Hosokawa et al., 2019). We used published CLIP-seq dataset from embryonic mouse brain to assess whether these aberrantly retained introns are SFPQ targets (Hosokawa et al., 2019). As illustrated in Ptk7 sequence, the retained intron 1 exhibited a much higher density of SFPQ binding peaks when compared to the constitutively spliced intron 6 within the same gene (Figure-9D). We further conducted analysis for all introns at the genomic level by comparing SFPQ binding peaks among introns with higher retention to introns with unaltered or lower retention. The comparison demonstrated that introns with elevated retention in Sfpq-depleted CNCCs showed significantly higher enrichment of SFPQ binding peaks (Figure-9E). We also analyzed SFPQ CLIP-seq peaks at the gene level, comparing genes with elevated IR to genes with no IR changes and showed that genes containing elevated IR in Sfpq-depleted CNCCs exhibited a trend towards higher enrichment of SFPQ binding peaks based on this embryonic brain dataset (Figure-9F). These data suggest that introns with aberrant IR elevation in Sfpq-depleted CNCCs are splicing targets bound by SFPQ.

To understand how SFPQ targets these matrix, Wnt and neuronal genes and introns, we examined molecular characteristics shared among SFPQ-regulated genes and recognized that they are long genes (>100kd in length) and retain long introns (>10kd in length). Of note, Wnt pathway gene Wwox and matrix genes St6galnac3 are 913kd and 526kd in length, and their retained introns are 639kd and 215kd in length, respectively, in contrast to a median size of 0.6 ∼ 2.4kb for introns in the mouse genome (Hong, Scofield, & Lynch, 2006). In the human and mouse genome, long genes mostly occur because of long introns (Breschi, Gingeras, & Guigó, 2017; Lopes, Altab, Raina, & de Magalhães, 2021). We then plotted the length of SFPQ-regulated genes against genes across the genome and demonstrated that the genes with increased IR in Sfpq-depleted CNCCs have a median length of ∼100kb. This is much longer than gene length in the control groups, where the medial length is around 10kb (Figure-9G). These findings suggest that SFPQ-regulated splicing facilitates regulation of long gene expression during craniofacial development.

SFPQ, EWSR1, TAF15 and TRA2B regulates distinct transcriptional and splicing events

To assess genes and IR co-regulated by SFPQ and PRMT1, we cross-analyzed genes with intron retention in both Sfpq-depleted CNCCs and Prmt1-deficient CNCCs. In Sfpq-depleted CNCCs, 179 genes exhibited significantly higher IR in both siRNA groups. In Prmt1-deficient CNCCs, 773 genes exhibited significantly higher IR in all embryos analyzed (Figure-10A). The cross-analysis showed that SFPQ-regulated intron retention bears partial (64 out of 773) overlap with PRMT1-regulated introns, exemplified by matrix genes Col4a2, Adam12, Ntn1, App, St6galnac3, Galnt11 Galnt10, and Asph (Figure-10A-E). Wnt signaling genes Wwox and Dkks, and neuronal genes that regulate membrane potential (38 out of 303 genes) was also identified among genes with decreased expression in Prmt1 CKO CNCCs, suggesting the regulation of these genes by the PRMT1-SFPQ pathway. The overlap may be under-represented because SFPQ was reduced to 50% by the depletion, but the partial overlay does suggest a notion that PRMT1 regulates intron retention through multiple splicing factors.

SFPQ, EWSR1, TAF15, and TRA2B regulate distinct transcriptional and splicing programs.

(A) Genes with increased IR events in SFPQ-depleted CNCCs demonstrated 8.28% (64 out of 773) overlap with Prmt1 deficient CNCCs. (B-E) Matrix genes Col4a2, Adam12, Ntn1, App, St6Galnac3, Galnt10 and Asph were regulated by both Prmt1 deletion and SFPQ depletion. Bar graph showing elevated IR (B, C) and reduced mRNA abundance (D, E) in CNCCs. TPM, transcripts per million. * p<0.05 siSFPQ vs siControl. * p<0.05 Prmt1 CKO vs Control. Control: Wnt1-Cre; R26RtdTomato. Prmt1 CKO: Wnt1-Cre; Prmt1fl/fl; R26RtdTomato. (F-S) ST2 cells were transfected with control or SFPQ, EWSR1, TAF15, TRA2B siRNAs and mature mRNA was extracted for sequencing. GSEA demonstrated enrichment of ECM genes among downregulated genes upon depletion of SFPQ, EWSR1, TRA2B, and TAF15 (F-I). Pie charts showed the percentage of genes with increased (dark grey) or decreased IRI (light grey) upon SFPQ, EWSR1, TAF15 and TRA2B deletion (J-M). rMATS analysis showed widespread and significant splicing changes following depletion of SFPQ, EWSR1, TAF15, and TRA2B (N). Postn mRNA expression was decreased in all four depletion groups (O). Exon skipping events of Postn were noted in all four depletion groups, illustrated by Sashimi plots (P) and quantified by bar graphs based on rMATS analysis (Q-S). * p<0.05 siRNAs vs siControl. siControl, control siRNA. siSFPQ#1 and siSFPQ#2, two independent SFPQ siRNAs.

We further characterized gene expression and intron retention regulated by SFPQ, EWSR1, TAF15 and TRA2B using ST2 cells. To this end, cells were transfected with control siRNA or two independent siRNAs targeting each splicing factor, and then poly(A)+ mRNA was purified followed by sequencing. GSEA analysis demonstrated that depletion of SFPQ, EWSR1, TRA2B, and TAF15 each caused downregulation of ECM genes (Figure-10F-I), suggesting that the four splicing factors all promote ECM gene expression. Global analysis for differentially expressed genes further showed that these splicing regulators control distinct biological functions (Supplemental Figure-S6). We also analyzed intron retention following their knockdown. siSFPQ caused differential IR in 514 genes, with increased IR in 171 genes and decreased IR in 370 genes (Figure-10J, Supplemental Table S4). siEWSR1 caused increased IR in 58 genes and decreased IR in 318 genes (Figure-10K, Supplemental Table S4), where genes with elevated IR belonged to diverse biological functions (Supplemental Table S5). In contrast, siTAF15 and siTRA2B caused predominantly decrease in IR with IR increase in <3% of genes with differential IR (Figure-10L-M, Supplemental Table S4). These findings suggest that the roles of EWSR1, TAF15 and TRA2B are distinct from SFPQ in the regulation of IR. Bioinformatic analysis for EWSR1, TAF15 and TRA2B didn’t reveal discernable overlap between ECM genes and IR changes, suggesting that they regulate ECM gene expression through IR-independent mechanisms in ST2 cells. Examination of other types of alternative splicing events by rMATS analysis (multivariant analysis of transcript splicing, MATS) revealed many splicing changes (Figure-10N, Supplemental Table S6). For example, the ECM gene Postn is known to exhibit exon skipping at exon 17 and 21, which affects Postn expression and activity that alter mandibular morphogenesis (Ishihara et al., 2023). We observed that the mRNA expression of Postn is reduced in all four splicing factor depletion groups and exon skipping of Postn showed a trend towards enhancement by the depletion, as illustrated by Sashimi plots and quantified from rMATS analysis (Figure-10O-10S).

To identify additional splicing regulators accountable for IR changes in Prmt1 deficient embryos, we conducted in-silico analysis using RNA-seq data of mandibular CNCCs from control and Prmt1 CKO embryos by performing rMAPS2 (RNA map analysis) subsequent to rMATS (Hwang et al., 2020; Park, Jung, Rouchka, Tseng, & Xing, 2016; Y. Wang et al., 2024). A list of RNA-binding proteins (RBPs) was identified with highly significant binding scores in the differentially expressed genes, suggesting them as potential downstream mediators for PRMT1-regulated splicing changes. This in-silico approach predicted additional splicing regulators that may function downstream of PRMT1 in elevating intron retention, including SRSF9, Lin28A, PABPN1, ESRP2, TARDBP, KHDRBS1 and FXR1 (Supplemental Table S7). SFPQ is also among the RBPs with top occurrences, with SFPQ binding motifs depicted in the vicinity of alternative splicing events (Supplemental Figure-S7A-S7E, Supplemental Table S8). Taken together, PRMT1-regulated splicing in CNCCs is mediated by a plexus of splicing factors including SFPQ.

Discussion

In this study, we revealed prevalent intron retention within CNCCs during embryonic development and demonstrated that PRMT1-SFPQ pathway regulates the splicing of matrix and Wnt pathway genes during craniofacial development. Genetic deletion of Prmt1 or depletion of Sfpq in CNCCs caused aberrant IR that triggers NMD to degrade intron-retaining transcripts. To our knowledge, this study is the first to characterize functional significance of IR-induced NMD in CNCCs and during craniofacial development. A main feature shared by SFPQ-regulated genes in CNCCs is gene length, with a median length of 100kb, in contrast to the median length of ∼10kb across the genome. Introns exhibiting high retention also tend to be long introns, for example, intron 4 of Col4a2 which displayed dramatically increased retention in Sfpq deficiency boasts a length of 45kb. Long genes with long introns are linked to neuronal function, where retained introns have been proposed to facilitate a rapid response to external stimuli, allowing a time frame shorter than that required for de novo transcription (Mauger, Lemoine, & Scheiffele, 2016; Ni et al., 2016).

Genes over 100kb in length require two to several hours for transcription based on a rate of 1-2kb/minute, which poses a challenge for the fast tempo of embryogenesis. This intron-regulated mechanism may represent a physiological strategy that enables rapid production of matrix proteins when they are required for craniofacial morphogenesis. This post-transcriptional regulation allows cells to control the rate of transcript production, and to achieve rapid protein translation without waiting for the time period required for transcription. Long genes regulated by this mechanism are highly enriched in matrix formation, responsible for the production of building blocks for bone, cartilage, and connective tissues. Since spliceosomopathies preferentially affect the craniofacial skeleton and limb, IR-regulated matrix expression provides mechanistic insights for the higher susceptibility of these tissue types to spliceosome dysfunction. Many of the PRMT1-SFPQ pathway regulated matrix and Wnt signaling genes are associated with congenital defects. For example, in our study, the long gene Wwox exhibited increased retention of long introns (introns 3 and 4) and decreased expression upon Prmt1 or Sfpq deficiency. In human patients, pathogenic variants of WWOX with large deletion within the long introns have been associated with epileptic encephalopathy syndrome manifesting shared facial phenotype (Dvinge & Bradley, 2015). The deletion within these long introns spanning the SFPQ binding region may affect SFPQ recognition and subsequent splicing of WWOX transcripts. Variants of another SFPQ target, Ptk7, are associated with neural tube defects and scoliosis (Berger, Wodarz, & Borchers, 2017). SFPQ itself is also linked to neurodegenerative diseases including amyotrophic lateral sclerosis, frontotemporal dementia, and Alzheimer’s disease, and was recently identified as a genetic regulator in CNCCs through an integrated genomic analysis (Feng et al., 2021). Besides SFPQ-regulated targets, splicing factors we identified in the unbiased in-silico analysis, such as Rbfox2 and HuR (Supplemental Table S5), are also associated with craniofacial defects in human patients and mouse studies (Cibi et al., 2019; Glessner et al., 2014; Homsy et al., 2015; McKean et al., 2016; Verma et al., 2022). We propose that IR-induced downregulation acts as a causal mechanism for dysregulated gene expression that contributes to craniofacial deformity.

IR-triggered NMD in controlling the expression of matrix genes in CNCCs presented a distinct mechanism in CNCCs. NMD is the RNA turnover pathway to degrade aberrant RNAs, driven by a protein complex composed of SMG and UPF proteins(Han et al., 2018; Tan, Stupack, & Wilkinson, 2022). Intron retention-triggered NMD in cancer has both promotive and suppressive roles and NMD inhibitors has been tested for cancer therapy and recently cancer immunotherapy(Tan et al., 2022). During embryonic development, the functional significance of NMD machinery is implied by human genetic findings and mouse genetic models. SMG9 mutation in human patients causes malformation in the face, hand, heart and brain (Shaheen et al., 2016). Smg6, Upf1, Upf2 and Upf3a knockout mouse die at early embryonic stages (E5.5-E9.5), and Smg1 gene trap mutant mice die at E12.5, all presenting severe developmental defects (Han et al., 2018). In this study, upon Prmt1 or Sfpq deficiency, IR-induced NMD adopted a pathogenic role, degrading aberrant intron-retaining transcript to reduce gene expression. We also demonstrated a physiological role for IR-triggered NMD that tunes matrix gene expression during embryogenesis, as inhibition of NMD using chemical inhibitors led to accumulation of transcripts such as Alpl, Eln and Loxl1.

In mammalian cells, PRMT1 methylates over forty splicing factors to modulate protein stability, subcellular localization, thereby influencing their accessibility to mRNA (Thandapani et al., 2013; Wu et al., 2022). PRMT1-catalyzed arginine methylation also affects their RNA binding affinity and specificity, altering their activity and the splicing product (Fedoriw et al., 2019; Fong et al., 2019). Therefore, it is not surprising that the disruption of this methyltransferase leads to splicing defects. We recognized SFPQ as a key PRMT1 substrate in CNCCs and demonstrated abundant SFPQ methylation in the mandibular and maxillary primordium. Our previous work also revealed SFPQ methylation on arginine (Arg) 7 and 9 depends on PRMT1(Hartel et al., 2019). This aligns with previous work demonstrating that PRMT1 is responsible for the association of SFPQ with mRNA in mRNP complexes in mammalian cells (Snijders et al., 2015). Our findings on altered neuronal genes in Sfpq-depleted CNCCs also echoed previous reports in which SFPQ regulates many neuronal genes involved in axon extension, branching, viability, and synaptogenesis via direct binding to their intronic regions (Cosker, Fenstermacher, Pazyra-Murphy, Elliott, & Segal, 2016; Ruskin, Zamore, & Green, 1988). Loss of SFPQ function compromises splicing patterns, especially the accurate splicing of long introns in embryonic brain and AML samples (Luisier et al., 2018; Taylor et al., 2022; Thomas-Jinu et al., 2017). Findings in our study further revealed previously unappreciated roles of SFPQ in the regulation of matrix genes through IR, which bear functional significance in bone and cartilage matrix deposition and craniofacial skeleton formation. Manipulation of SFPQ expression (knockdown with 50% efficiency) demonstrated that SFPQ shares around a partial overlap with PRMT1 in IR-regulated neural crest gene expression. Given that PRMT1 methylates over 40 splicing factors in mammals and many of these splicing factors are abundantly expressed in embryonic CNCCs, we expect that PRMT1-regulated intron retention will be mediated by multiple splicing regulators besides SFPQ. This notion is further supported by the identification of additional splicing regulators via in-silico analysis. Additionally, PRMT1-governed splicing will be mediated by multiple RNA binding protein encompassing all types of alternative splicing, as suggested by PRMT1-mediated methylation of EWSR1 and TRA2B, and the distinct splicing footprints among EWSR1, TAF15, TRA2B and SFPQ.

Altogether, these findings demonstrate that the PRMT1-SFPQ pathway regulates IR in CNCCs and suggests IR-triggered NMD as a mechanism that controls matrix and Wnt signaling gene expression during craniofacial morphogenesis.

Material and methods

Animals

Prmt1fl/fl mice were generously provided by Dr. Stephane Richard (McGill University). Wnt1-Cre; Prmt1fl/fl; R26R tdTomato and Wnt1-Cre; R26RtdTomato mice were obtained by mating Wnt1-Cre, R26RtdTomato mice (Jax #009107 and #007914) with Prmt1fl/fl mice. Animals were genotyped using established protocols (Chai et al., 2000; Yu et al., 2009) and all care and experiments followed USC’s IACUC protocols.

Proximity Ligation Assay (PLA)

PLA was performed using a Duolink In Situ kit (Millipore Sigma Cat# DUO92101-1KT), SFPQ antibody (Everest Biotech Cat#EB09523), and pan-methyl arginine antibody, Asymmetric Di-Methyl Arginine (ADMA) (Cell Signaling Technology Cat# 13522) to detect methyl-SFPQ. Tissue sections were blocked with the Duolink blocking solution in a humidity chamber for 1 hour at 37°C before incubating with the primary antibodies (diluted 1:50 in Duolink PLA probe diluent) overnight at 4°C. The next day, tissue sections were incubated with the Goat-PLUS and Rabbit-MINUS PLA probes for 1 hour at 37°C, followed by ligase and amplification solution. After the final washes, tissue sections were mounted with Duolink In Situ mounting media with DAPI to counterstain nuclei. Images were visualized under the Confocal microscope. The number of PLA signals that appear as punctuate green signals in the nuclei of the cells and the quantification was performed with CellProfille software.

Tissue Processing and Immunofluorescence Staining

E13.5 control and Wnt1-Cre; Prmt1fl/fl mice embryos were fixed in 4% PFA overnight, dehydrated, embedded in O.C.T., sectioned at 8 µm thickness in sagittal orientation, and mounted on glass slides. The samples were submitted to antigen retrieval using Antigen Unmasking Solutions (Vector Laboratories Cat# H-3300-250) and washed and incubated in 0.5% Triton X-100 in PBS for 20 minutes for permeabilization. The samples were briefly washed with PBS, blocked with 10% goat serum for 1 hour, and then incubated with primary antibody SFPQ (Cell Signaling Technology Cat#23020, 1:100) at 4°C overnight in a moisture chamber. After this, the samples were blocked with a secondary antibody (1:400, goat anti-rabbit IgG (H+L) Alexa Fluor 488 Cat# A-11008) for 1 hour followed by DAPI diluted 1:1000 in PBS for 10 minutes. Finally, the samples were mounted with mounting media (Electron Microscopy Science Cat# 1798510) and covered with cover glass for imaging using Keyence BZ-X800 and Leica DMI 3000B microscopes and quantification with CellProfille software.

Western Blotting

Heads of E13.5 control and Wnt1-Cre; Prmt1fl/fl mice were dissected, followed by removal of the brain. The tissue was homogenized with a pestle followed by lysis with RIPA Lysis and Extraction Buffer. The total protein concentration was determined by comparison with BSA standards. Twenty micrograms of total protein from each sample were loaded into each well of a 10% polyacrylamide gel. Western blot analysis was carried out as previously described (Iwata et al., 2012). Antibodies against PRMT1 (Cell Signaling, Cat# 2449; 1:1000), SFPQ (Cell Signaling Technology Cat#23020, 1:1000), and GAPDH (Cell Signaling, Cat# 97166; 1:5000) were used for Western blot. The samples were analyzed three times in independent blots and the bands were quantified by optical densitometry. The analysis was performed using ImageJ digital imaging processing software (ImageJ 1.48v, National Institutes of Health, Bethesda, MD, USA). The expression of each analyzed protein was normalized with GAPDH.

NCC collection and FACS analysis

The heads of E13.5 Wnt1-Cre; R26RtdTomato and Wnt1-Cre; Prmt1fl/fl;R26RtdTomato mice were dissected and placed in a sterile tube containing Hanks’ Balanced Salt Solution (Thermo Fisher Scientific Cat# 14175079). Following centrifugation, the supernatant was discarded, and the tissue was subjected to TrypLE (Thermo Fisher Scientific Cat# 12605010) incubation for 5-15 minutes at 37°C on a rotator. The resulting dissociated tissue was neutralized by the addition of fetal bovine serum (FBS) and then filtered through a 40μm strainer to obtain a single-cell suspension. After centrifugation, cells were resuspended in an appropriate volume of serum-free medium for FACS analysis. Sorting was performed based on tdTomato fluorescence (Excitation/emission: 554/582 nm), and td-positive cells were collected into separate tubes containing DMEM (Genesee Scientific Cat#25-501) supplemented with 20% FBS.

Primary CNCC culture and siRNA transfection

CNCCs were cultured in DMEM supplemented with 20% FBS in an incubator until they attach to the bottom. Then, reverse transfection with control siRNA (Qiagen Cat# 1027310), SFPQ siRNA #1 (Qiagen Cat#SI05783848), and SFPQ siRNA #2 (Qiagen Cat#SI05783876) at 40 nM using Lipofectamine RNAiMax transfection reagent (Invitrogen) for siRNA delivery were performed. 48h later, total RNA was isolated using TRIzol reagent (Invitrogen) following the manufacturer’s protocols followed by mRNA isolation using NEBNext High-Input Poly(A) mRNA Magnetic Isolation Module (NEBNext® Cat#E3370S).

mRNA isolation and sequencing

Poly(A) mRNA isolation was extracted from total RNA by using the NEBNext High-Input Poly(A) mRNA Magnetic Isolation Module (NEBNext® Cat#E3370S). Five sets of primary isolated CNCCs from control (Wnt1-Cre;R26R tdTomato) or Prmt1 CKO (Wnt1-Cre; Prmt1fl/fl; R26R tdTomato), and three sets of isolated CNCCs transfected with siControl or siSFPQs were sequenced at 40 million reads sequencing depth and 150 bp paired-end sequencing.

Bioinformatic Analysis of Differentially Regulated Genes (DEG) and differential Intron Retention (IR)

The sequencing quality of RNA-seq libraries was assessed by FastQC v0.11.8 (https://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Low-quality bases and adapters were trimmed using Trim Galore (v0.6.7,https://www.bioinformatics.babraham.ac.uk/projects/trim_galore/). The reads were mapped to mouse genome mm10 using hisat2 (v2.1.0) (Kim, Paggi, Park, Bennett, & Salzberg, 2019). The mapped sam files from hisat2 were converted to bam files which were then further turned to sorted bam files by samtools (v1.7) (H. Li et al., 2009). Mapped reads were then processed by htseq-count (v1.99.2) to calculate the read count of all genes (Anders, Pyl, & Huber, 2015). The sorted bam files were then given to IRTools (https://github.com/WeiqunPengLab/IRTools/) to calculate the IR of each gene and the read count of genes’ constitutive intronic regions (CIRs) and constitutive exonic regions (CERs). The expression level of a gene was expressed as TPM (Transcripts Per Kilobase Million) value. EdgeR was used to identify differentially expressed genes (DEGs) by requiring ≥ 1.5-fold expression changes and adjusted p-value < 0.05 (Robinson, McCarthy, & Smyth, 2010). The intron retention index (IRI) of a gene is defined as the ratio of the overall read density of its CIRs to the overall read density of its CERs. Genes with 0< IRI <1 and CER expression greater than 1 were selected to identify differential IR genes. Student’s t-test was used to identify differential IR genes by requiring ≥ 1.5-fold log2(IR) changes and p-value < 0.05. Gene Ontology analysis was done by DAVID (https://david.ncifcrf.gov/tools.jsp) and Metascape (https://metascape.org/gp/index.html#/main/step1) (Sherman et al., 2022; Zhou et al., 2019).

The RNA-seq data are deposited to the Gene Expression Omnibus (GEO) database with accession number GSE266474.

rMATS and rMAPS Analysis

rMATS (https://github.com/Xinglab/rmats-turbo) and rMAPS2 (http://rmaps.cecsresearch.org/) were used to find potential RNA binding Proteins (RBPs) that contribute to differential splicing events (Hwang et al., 2020; Y. Wang et al., 2024). rMATS was first used to identify differential alternative splicing events between wild type and PRMT1 CKO samples by requiring FDR < 0.05. The output of rMATs were then fed to rMAPS2 to identify significant RBPs contributing to each type of splicing events.

Enrichment analysis of SFPQ binding

To assess the SFPQ binding of the genes whose IR is regulated by SFPQ, we downloaded SFPQ CLIP-seq peak data from GEO (GSE96080). The CLIP-seq peaks shared between the replicates were filtered by p < 0.01 and log(FC) > 1 to obtain the significant peaks for downstream analysis. Next, we defined the SFPQ-regulated IR up and down gene sets by combining those from two SFPQ knockdown conditions respectively. We constructed the control gene set by selecting those with no significant change in IR (IR difference < 0.0005, 0.9 < IR FC < 1.1, and p > 0.5) in either SFPQ knockdown conditions. For each gene set, the percentage of genes with introns overlapping SFPQ CLIP-seq peaks was calculated. Furthermore, for each gene we evaluated the SFPQ peak density as the number of SFPQ peaks on the gene divided by the gene length.

Semi-quantitative and Quantitative PCR

For both experiments, Control, and Wnt1-Cre; Prmt1fl/fl mandibles, and siRNA-transfected CNCCs, mRNAs were quantified by real-time PCR with IQ Sybr Green Supermix (Bio-Rad) and normalized against GAPDH mRNA levels. Relative changes in expression were calculated using the ΛλΛλCt method. Primer sequences are listed in Supplemental Table S9.

Cleavage Under Target and Tagmentation (CUT&Tag)

CUT&Tag was adapted from in Epicypher DIY protocol and Kaya-Okur, H.S., Janssens, D.H., Henikoff, J.G. et al. Efficient low-cost chromatin profiling with CUT&Tag. Nat Protoc 15, 3264–3283 (2020). https://doi.org/10.1038/s41596-020-0373-x. Before extracted nuclei, 10 µL of ConA beads was resuspend twice in Bead Activation Buffer (20mM HEPES, pH7.9; 10mM KCl; 1mM CaCl2; 1mM MnCl2). 100,000 cells were washed once with PBS and and nuclei were extracted by resuspending cells in nuclear extraction (NE) buffer (20mM HEPES-KOH, pH 7.9, 20mM KCl, 0.1% Triton X-100, 20% Glycerol, 0.5mM Spermidine, and protease inhibitor) and incubating for 10 minutes on ice. Nuclei were collected by centrifuge at 600g for 3 minutes and resuspended in 100 µL of NE buffer per 100,000 nuclei. 100 µL of nuclei were transferred to PCR tube containing 10 µL of activated ConA beads, gently vortex, and incubate at room temp for 10 minutes. The supernatant was discarded and 50 µL of Antibody buffer (20mM HEPES, pH7.5; 150mM NaCl; 0.5mM Spermidine; 0.01% Digitonin; 2mM ETDA; and protease inhibitor) was added. An antibody against RNA Pol subunit 1 (Rbp1 CTD (4H8) mouse mAbm Cell Signal Tech.) was added and incubated on a nutator at 4C overnight. Beads was then incubated in 50µL of 10 µg/ml of secondary antibody was made in Digitonin150 buffer (20mM HEPES, pH7.5; 150mM NaCl; 0.5mM Spermidine; 0.01% Digitonin; and protease inhibitor) at room temperature for 30 minutes on a nutator. Wash twice with Digitonin150 buffer before resuspending beads in pAG-Tn5 diluted in Digitonin300 buffer (20mM HEPES, pH7.5; 300mM NaCl; 0.5mM Spermidine; 0.01% Digitonin; and protease inhibitor) and incubating for 1 hour at room temperature on a nutator. Wash twice with Digitonin300 buffer then start the tagmentation reaction by resuspending beads in Tagmentation buffer (Digitonin300 supplemented with 10mM MgCl2) and incubated at 37C in thermocycler. Beads were washed once with 50µL TAPS Buffer (10mM TAPS pH 8.5; 0.2mM EDTA) then incubated in 50µL of Release Buffer (10mM TAPS pH 8.5; 0.1% SDS; 0.4 mg/µL Proteinase K; 15mM EDTA) at 55C for 1 hour; 75C for 20 minutes; and then cool down to 20C. Tagmented DNA were purified with 1.8x AMPureXP beads and eluted with 20µL of DNase/RNase free H2O. The library was prepared with dual index primers and NEBNext High-Fidelity 2X PCR master mix (M0541) and amplified for 14 cycles. Library DNA was purified with 1.3x AMPureXP beads and eluted with 15µL 0.1xTE buffer. Sequencing run was performed on the NovaSeq X Series platform with 10B reagent kit (300 Cycle). NovaSeq Control Software 1.2.2.48004 was used for sequencing and Illumina BCL Convert v4.2.7 to convert base call (BCL) files into FASTQ files.

Statistical analysis

Two-tailed Student’s t-tests or Fisher’s exact tests were applied for statistical analysis. For all graphs, error bars represent standard deviations. A P-value of <0.05 was considered statistically significant.